---
title: "Complete methodology "
author: "María Garcés Blázquez (100453214) - Master in Computational Social Science"
date: 'UC3M 2025--2026'
format:
  html:
    theme: flatly
    highlight-style: tango
    number-sections: false
    toc: true
    toc-depth: 3
    toc-location: left
    self-contained: true
execute:
  echo: true
  warning: false
  message: false
---

## Overview

This document contains the complete practical application of the methodology 
proposed in the Master's Thesis (TFM) for the estimation and prediction of 
energy poverty at the census section level in the municipality of Madrid.

The analysis is structured in three sequential phases that integrate 
socioeconomic data from the Living Conditions Survey (ECV), geospatial 
information from the Income Distribution Atlas (INE), and environmental 
variables derived from satellite observation (Google Earth Engine):

**Phase 1 — Construction of the Energy Vulnerability Index (EVI)**  
Definition and calculation of the Energy Vulnerability Index based on 
expenditure, income, and household conditions available in the ECV. 
Imputation of missing values using missForest and cross-validation with 
alternative methods (Amelia, MICE, quintile median).

**Phase 2 — Small Area Estimation (SAE)**  
Application of the Synthetic Regression Estimator (Rao & Molina, 2015) 
using a logistic mixed-effects model to predict the EVI in the 2,443 
census sections of Madrid. Justification of the EVI > 15 threshold 
through triple validation: Eurostat criterion, empirical distribution, 
and comparison with independent indicators from the INE Atlas.

**Phase 3 — Predictive Models**  
Construction of two complementary Random Forest models:  
(i) a socioeconomic model combining census and geospatial variables 
to classify sections according to energy poverty;  
(ii) a physical vulnerability model based exclusively on environmental 
variables (Google Earth Engine) that identifies sections with adverse 
physical conditions (thermal, vegetation, building characteristics).

**Phase 4 — Discrepancy Analysis and Group Characterization**  
Identification and characterization of four groups: concordant, 
non-vulnerable, spatial false positives, and spatial false negatives. 
Analysis of the differential characteristics of each group and 
statistical validation of the observed differences.

Each phase includes the R code used, the outputs generated (tables, 
plots, and result files), and the interpretation of the findings. 
The ultimate goal is to provide a reproducible tool for the territorial 
identification of energy poverty that combines traditional data sources 
with remote sensing.


## Phase 1: Construction EVI

Phase 1 builds the analytical foundation in two sequential steps.
First, a master census file is constructed by integrating eight administrative data sources at the census section level (CUSEC): income, demographic, inequality, and poverty threshold indicators from the INE Atlas of Income Distribution 2021; household composition and nationality data from the Padrón Municipal 2022; population counts from the INE Population Census 2021; and building age and construction year data from the CAM Building Census 2022–2024. The resulting dataset covers all 2,443 census sections of Madrid with approximately 30 variables per section.
Second, missing values concentrated in poverty threshold indicators (304 sections, 12.4%) are addressed through a systematic comparison of five imputation methods — missForest, MICE, Amelia, conditional mean, and conditional median by income quintile — evaluated via 10-fold cross-validation with 20% of values artificially masked. missForest is selected as the optimal method based on its lowest Mean Absolute Error (0.95 percentage points, 28% below MICE at 1.32 pp) and absence of systematic bias in the error distribution. The final imputed dataset contains zero missing values across all retained variables and serves as the input for the SAE downscaling in Phase 2.

#### Methodology Summary

This analysis comprises two main phases:

**Part 1: Data Integration and Imputation**

- Construction of a master census file integrating 10 data sources (Atlas Nacional, 
  Padrón Municipal, INE Population, and CAM Census)
- Identification and treatment of missing values in key auxiliary variables
- Comparative evaluation of 5 imputation methods (missForest, MICE, Amelia, 
  mean, and median) using cross-validation
- Selection of missForest as the final imputation method based on RMSE performance

**Part 2: Small Area Estimation**

- Bidimensional stratification of census sections by income (5 quintiles) and 
  demographic vulnerability (4 quartiles), resulting in 20 homogeneous strata
- Classification of ECV households into the same stratification scheme
- Probabilistic matching of ECV households to census sections within the same stratum
- Direct weighted estimation of EVI by section, with fallback imputation using 
  stratum means for sections without assigned households

#### Key Results

- **Coverage**: 100% (2,443/2,443 sections)
- **Stratification quality**: R² = 0.911 (income), R² = 0.857 (demographics)
- **Direct estimation**: 42.2% of sections (1,032) with EVI estimated from 
  matched households
- **Stratum imputation**: 57.8% of sections (1,411) with EVI imputed from 
  stratum means
- **Average EVI**: 18.1% (range: 0.0% - 100.0%)

This SAE-derived EVI serves as the baseline for Phase 3, where Random Forest 
will integrate physical variables (building age, land surface temperature, NDVI) 
to identify spatial discrepancies and detect "invisible" energy poverty not 
captured by survey-based approaches.

#### Libraries

```{r}
library(readxl)
library(mice)
library(missForest)
library(Amelia)
library(ggplot2)
library(Hmisc)
library(patchwork)
library(lme4)
library(MuMIn)
library(tidyverse)
library(sf)
library(flextable)
library(scales)
```

#### Paths

```{r}
# --- INE Atlas (Income Distribution Atlas) ---
path_atlas_income       <- "data/Atlas_renta.csv"
path_atlas_demo         <- "data/Atlas_demografia.csv"
path_atlas_inequality   <- "data/Atlas_desigualdad.csv"
path_atlas_threshold_sx <- "data/Atlas_umbral_sexo.csv"

# --- Municipal Register and Population Data ---
path_register   <- "data/2022.xlsx"
path_population <- "data/INE_Madrid_Poblacion_Secciones_2021_2025.xlsx"

# --- CAM Census (Community of Madrid Building Census) ---
path_census22 <- "data/dataset_22_censal.xlsx"
path_census24 <- "data/dataset_24_censal.xlsx"


gee_path <- "data/Madrid_Urban_Full_Audit_Stats_FIXED.csv"
```


#### CUSEC Extraction Function

Atlas INE data sources include a text field with census section identifiers 
embedded in descriptive strings (e.g., "2807901001 Madrid sección 01001"). 
This helper function extracts the 10-digit CUSEC code from such strings:

1. `str_trim(x)` removes leading/trailing whitespace
2. `str_extract(..., "^[0-9]{10}")` captures the first 10 consecutive digits
3. Returns the clean CUSEC identifier (e.g., "2807901001")

The CUSEC format is: `PPMMMDDSSS` where:
- PP = Province code (28 = Madrid)
- MMM = Municipality code (079 = Madrid capital) 
- DD = District code
- SSS = Section code

```{r}
extract_cusec_atlas <- function(x) {
  str_extract(str_trim(x), "^[0-9]{10}")
}
```


### Part 1: Census Master File Construction

This section integrates 10 administrative data sources at the census section 
level (CUSEC) to create a unified dataset with socioeconomic, demographic, 
and housing variables for all 2,443 sections in Madrid.

Data sources:

- Atlas Nacional (Income, Demographics, Inequality, Poverty Thresholds)
- Padrón Municipal 2022 (Household Composition, Nationality)
- INE Population Census 2021
- CAM Census 2022-2024 (Building Age, Construction Year)

#### Demographics
```{r}
demo <- read_delim(path_atlas_demo, delim = ";", locale = locale(decimal_mark = ","), 
                   show_col_types = FALSE) |> 
  # Rename original Spanish columns from INE to English
  rename(municipalities = Municipios,
         sections = Secciones,
         period = Periodo,
         total = Total,
         demographic_indicators = `Indicadores demográficos`) |> 
  # Filter: Madrid municipality code (28079) and Year 2021
  filter(str_detect(municipalities, "^28079"), str_trim(sections) != "", period == "2021") |> 
  # Extract 10-digit CUSEC and parse values
  mutate(CUSEC = extract_cusec_atlas(sections),
         value = as.numeric(str_replace_all(total, ",", "."))) |> 
  filter(!is.na(CUSEC)) |> 
  select(CUSEC, indicator = demographic_indicators, value) |> 
  # Select key demographic indicators (filter still uses Spanish strings because
  # those are the actual VALUES inside the cells, not column names)
  filter(indicator %in% c(
    "Tamaño medio del hogar",                   # Average household size
    "Porcentaje de población menor de 18 años", # % population under 18
    "Porcentaje de población de 65 y más años", # % population 65+
    "Porcentaje de hogares unipersonales",      # % single-person households
    "Porcentaje de población española"          # % Spanish population
  )) |> 
  # Translate indicator VALUES from Spanish to English
  mutate(indicator = case_when(
    indicator == "Tamaño medio del hogar" ~ "mean_household_size",
    indicator == "Porcentaje de población menor de 18 años" ~ "pct_under_18",
    indicator == "Porcentaje de población de 65 y más años" ~ "pct_over_65",
    indicator == "Porcentaje de hogares unipersonales" ~ "pct_single_person",
    indicator == "Porcentaje de población española" ~ "pct_spanish"
  )) |> 
  # Reshape to wide format (column names will now be in English)
  pivot_wider(names_from = indicator, values_from = value) |> 
  # Calculate % foreign population
  mutate(pct_foreign = 100 - pct_spanish) |> 
  select(-pct_spanish)
```

#### Income
```{r}
income <- read_delim(path_atlas_income, delim = ";", locale = locale(decimal_mark = ","),
                     show_col_types = FALSE) |> 
  # Rename original Spanish columns from INE to English
  rename(municipalities = Municipios,
         sections = Secciones,
         period = Periodo,
         total = Total,
         income_indicators = `Indicadores de renta media y mediana`) |> 
  # Filter: Madrid municipality (28079) and Year 2021
  filter(str_detect(municipalities, "^28079"), str_trim(sections) != "", period == "2021") |> 
  # Extract CUSEC and parse income values (remove thousands separator)
  mutate(CUSEC = extract_cusec_atlas(sections),
         value = as.numeric(str_replace_all(total, "\\.", ""))) |> 
  filter(!is.na(CUSEC)) |> 
  select(CUSEC, indicator = income_indicators, value) |> 
  # Select key income indicators (Spanish strings because they are cell VALUES)
  filter(indicator %in% c(
    "Media de la renta por unidad de consumo",     # Mean income per equivalence unit
    "Mediana de la renta por unidad de consumo",   # Median income per equivalence unit
    "Renta neta media por hogar"                   # Mean net household income
  )) |> 
  # Translate indicator values to English (will become column names after pivot)
  mutate(indicator = case_when(
    indicator == "Media de la renta por unidad de consumo" ~ "mean_equiv_income",
    indicator == "Mediana de la renta por unidad de consumo" ~ "median_equiv_income",
    indicator == "Renta neta media por hogar" ~ "net_household_income"
  )) |> 
  # Reshape to wide format (column names will now be in English)
  pivot_wider(names_from = indicator, values_from = value)
```

#### Inequality
```{r}
inequality <- read_delim(path_atlas_inequality, delim = ";", locale = locale(decimal_mark = ","),
                         show_col_types = FALSE) |> 
  # Rename original Spanish columns from INE to English
  rename(municipalities = Municipios,
         sections = Secciones,
         period = Periodo,
         total = Total,
         inequality_indicators = `Índice de Gini y Distribución de la renta P80/P20`) |> 
  # Filter: Madrid (28079), Year 2021, and Gini coefficient indicator
  filter(str_detect(municipalities, "^28079"), str_trim(sections) != "", period == "2021",
         inequality_indicators == "Índice de Gini") |> 
  # Extract CUSEC and parse Gini values
  mutate(CUSEC = extract_cusec_atlas(sections),
         gini = as.numeric(str_replace_all(total, ",", "."))) |> 
  filter(!is.na(CUSEC)) |> 
  select(CUSEC, gini)
```

#### Poverty Thresholds
```{r}
threshold_sx <- read_delim(path_atlas_threshold_sx, delim = ";", locale = locale(decimal_mark = ","),
                           show_col_types = FALSE) |> 
  # Rename original Spanish columns from INE to English
  rename(municipalities = Municipios,
         sections = Secciones,
         period = Periodo,
         sex = Sexo,
         total = Total,
         income_distribution = `Distribución de la renta por unidad de consumo`) |> 
  # Filter: Madrid (28079), Year 2021, Total population (not sex-disaggregated)
  filter(str_detect(municipalities, "^28079"), str_trim(sections) != "", 
         period == "2021", sex == "Total") |> 
  # Extract CUSEC and parse threshold values
  mutate(CUSEC = extract_cusec_atlas(sections),
         value = as.numeric(str_replace_all(total, ",", "."))) |> 
  filter(!is.na(CUSEC)) |> 
  select(CUSEC, indicator = income_distribution, value) |> 
  # Select poverty threshold indicators (Spanish strings because they are cell VALUES)
  filter(indicator %in% c(
    "Población con ingresos por unidad de consumo por debajo 40% de la mediana",  # Below 40%
    "Población con ingresos por unidad de consumo por debajo 50% de la mediana",  # Below 50%
    "Población con ingresos por unidad de consumo por debajo 60% de la mediana",  # Below 60% (EU)
    "Población con ingresos por unidad de consumo por encima 140% de la mediana"  # Above 140%
  )) |> 
  # Translate indicator values to English (will become column names after pivot)
  mutate(indicator = case_when(
    indicator == "Población con ingresos por unidad de consumo por debajo 40% de la mediana" ~ "pct_below_threshold_40",
    indicator == "Población con ingresos por unidad de consumo por debajo 50% de la mediana" ~ "pct_below_threshold_50",
    indicator == "Población con ingresos por unidad de consumo por debajo 60% de la mediana" ~ "pct_below_threshold_60",
    indicator == "Población con ingresos por unidad de consumo por encima 140% de la mediana" ~ "pct_above_threshold_140"
  )) |> 
  # Reshape to wide format (column names will now be in English)
  pivot_wider(names_from = indicator, values_from = value)

cat("  Sections:", nrow(threshold_sx), "\n")
```

#### Household Composition

```{r}
register_composition <- read_excel(path_register, sheet = "Composicion del hogar", 
                                   skip = 4, col_names = FALSE) |> 
  # Filter: Valid 5-digit section codes
  filter(!is.na(...2), str_detect(as.character(...2), "^[0-9]{5}$")) |> 
  # Build CUSEC and extract household composition data
  mutate(
    section_code = as.character(...2),
    CUSEC = paste0("28079", str_sub(section_code, 1, 2), str_sub(section_code, 3, 5)),
    total_households = as.numeric(...3),
    # Total single-parent households (father + mother)
    n_single_parent = as.numeric(...8) + as.numeric(...9),
    pct_single_parent = (n_single_parent / total_households) * 100,
    # Female-headed single-parent households ONLY (column 9 = mother alone)
    n_single_mother = as.numeric(...9),
    pct_single_mother = (n_single_mother / total_households) * 100
  ) |> 
  filter(!is.na(total_households), total_households > 0) |> 
  select(CUSEC, total_households_register = total_households, 
         n_single_parent, pct_single_parent, n_single_mother, pct_single_mother)
```

#### Nationality

```{r}
register_nationality <- read_excel(path_register, sheet = "Tipo-Nacionalidad de sus miembr",
                                   skip = 4, col_names = FALSE) |> 
  # Filter: Valid 5-digit section codes
  filter(!is.na(...2), str_detect(as.character(...2), "^[0-9]{5}$")) |> 
  # Build CUSEC and calculate % all-foreign households
  mutate(
    section_code = as.character(...2),
    CUSEC = paste0("28079", str_sub(section_code, 1, 2), str_sub(section_code, 3, 5)),
    total_nationality = as.numeric(...3),
    pct_foreign_households = (as.numeric(...5) / total_nationality) * 100  # % all-foreign households
  ) |> 
  filter(!is.na(total_nationality), total_nationality > 0) |> 
  select(CUSEC, pct_foreign_households)

# Merge household composition and nationality data
register <- register_composition |> left_join(register_nationality, by = "CUSEC")
```

#### Population by Section
```{r}
population <- read_excel(path_population, sheet = 1, skip = 8, col_names = FALSE) |> 
  # Filter: Valid CUSEC codes for Madrid (28079)
  filter(!is.na(...1), str_detect(as.character(...1), "^28079[0-9]{5}")) |> 
  # Extract CUSEC and population data
  mutate(
    CUSEC = str_extract(as.character(...1), "^[0-9]{10}"),
    total_population = as.numeric(...6),
    foreign_population = as.numeric(...16),
    pct_foreign_pop = (foreign_population / total_population) * 100  # % foreign population
  ) |> 
  filter(!is.na(CUSEC), !is.na(total_population)) |> 
  select(CUSEC, total_population, pct_foreign_pop)
```

##### Building Age - Mean and Median (CAM Census)
```{r}
census24 <- read_excel(path_census24, sheet = "Sheet0", skip = 17,
                       col_names = c("name", "dwellings", "mean_surface", "median_surface", "mode_surface",
                                     "mean_building_age", "median_building_age", "mode_building_age")) |> 
  # Filter: Madrid district and section entries
  filter(str_detect(name, "^Madrid - Distrito [0-9]+ - Sección [0-9]+")) |> 
  # Build CUSEC from district and section numbers
  mutate(
    CUSEC = paste0("28079",
                   str_pad(str_extract(name, "(?<=Distrito )[0-9]+"), 2, "left", "0"),
                   str_pad(str_extract(name, "(?<=Sección )[0-9]+"), 3, "left", "0")),
    mean_building_age = as.numeric(mean_building_age),      # Mean building age (years)
    median_building_age = as.numeric(median_building_age)   # Median building age (years)
  ) |> 
  filter(!is.na(mean_building_age)) |> 
  select(CUSEC, mean_building_age, median_building_age)
```

#### Pre-1980 Buildings (CAM Census)
```{r}
census22 <- read_excel(path_census22, sheet = "Sheet0", skip = 11,
                       col_names = c("name", "total", "before_1900",
                                     "d1901_1920", "d1921_1940", "d1941_1950",
                                     "d1951_1960", "d1961_1970", "d1971_1980",
                                     "d1981_1990", "d1991_2000", "d2001_2005",
                                     "d2006_2010", "d2011_2015", "d2016_2020",
                                     "d2021_2025", "unknown")) |> 
  # Filter: Madrid district and section entries
  filter(str_detect(name, "^Madrid - Distrito [0-9]+ - Sección [0-9]+")) |> 
  # Build CUSEC and calculate % pre-1980 buildings (thermal inefficiency proxy)
  mutate(
    CUSEC = paste0("28079",
                   str_pad(str_extract(name, "(?<=Distrito )[0-9]+"), 2, "left", "0"),
                   str_pad(str_extract(name, "(?<=Sección )[0-9]+"), 3, "left", "0")),
    across(total:unknown, as.numeric),
    dwellings_pre1980 = before_1900 + d1901_1920 + d1921_1940 +
                        d1941_1950 + d1951_1960 + d1961_1970 + d1971_1980,
    pct_pre1980 = (dwellings_pre1980 / total) * 100  # % built before 1980 (NBE-CT-79)
  ) |> 
  filter(!is.na(total), total > 0) |> 
  select(CUSEC, total_dwellings = total, pct_pre1980)

cat("  Sections:", nrow(census22), "\n")
```

#### Join Full dataset

```{r}
master_census <- income |> 
  left_join(demo,         by = "CUSEC") |> 
  left_join(inequality,   by = "CUSEC") |> 
  left_join(threshold_sx, by = "CUSEC") |> 
  left_join(register,     by = "CUSEC") |> 
  left_join(population,   by = "CUSEC") |> 
  left_join(census24,     by = "CUSEC") |> 
  left_join(census22,     by = "CUSEC")

cat(sprintf("Master census: %d sections × %d variables\n", 
            nrow(master_census), ncol(master_census)))
```

#### Quality verification

##### Duplicate CUSECs

```{r}
n_dup <- master_census |> 
  group_by(CUSEC) |> 
  filter(n() > 1) |> 
  nrow()
n_dup
```
No duplicates, perfect.

##### Missing data

```{r}
master_census |> 
  summarise(across(everything(), ~sum(is.na(.)))) |> 
  pivot_longer(everything(), names_to = "variable", values_to = "n_na") |> 
  arrange(desc(n_na))
```

### Part 2: Missing Data Imputation

#### Overview

The master census file contains missing values in key variables required for 
stratification and modeling. This section addresses these gaps through 
systematic comparison of imputation methods.

**Missing data summary:**

- `pct_below_threshold_40/50/60`, `pct_above_threshold_140`: 354 NAs (14.5%)
- `net_household_income`, `median_equiv_income`: 50 NAs (2.0%)
- `pct_under_18`, `pct_over_65`, `mean_household_size`: 50 NAs (2.0%)

**Methods compared:**

1. **missForest** — Random Forest multivariate imputation
2. **MICE** — Multiple Imputation by Chained Equations (PMM)
3. **Amelia** — Expectation-Maximization with Bootstrap
4. **Conditional mean** — Mean by income quintile
5. **Conditional median** — Median by income quintile

**Validation strategy:**

- 10-fold cross-validation with 20% masked data
- Metric: RMSE (Root Mean Squared Error) on percentage variables
- Selection criterion: Lowest average RMSE across critical variables

**Missing data summary (after filtering by `mean_equiv_income`):**

- `pct_below_threshold_40/50/60`, `pct_above_threshold_140`: 304 NAs (12.4%)
- All demographic and housing variables: 0 NAs (filtered out with `mean_equiv_income`)

**Note:** 50 sections were excluded due to missing `mean_equiv_income` 
(stratification variable). Final analysis dataset: 2,443 sections.

#### Missing Data Diagnosis

```{r}
na_counts <- master_census |> 
  summarise(across(everything(), ~sum(is.na(.)))) |> 
  pivot_longer(everything(), names_to = "variable", values_to = "n_na") |> 
  mutate(pct_na = 100 * n_na / nrow(master_census)) |> 
  filter(n_na > 0) |> 
  arrange(desc(n_na))

na_counts
```

#### Method Comparison via Cross-Validation

```{r}
# Prepare data ONCE
census_ebp <-
  master_census |> 
  filter(!is.na(mean_equiv_income)) |> 
  mutate(income_quintile = ntile(mean_equiv_income, 5))

# Define variables ONCE
vars_to_impute <- c("pct_below_threshold_40", "pct_below_threshold_50", "pct_below_threshold_60",
                    "pct_above_threshold_140", "net_household_income", "median_equiv_income",
                    "pct_under_18", "pct_over_65", "mean_household_size",
                    "gini", "pct_single_parent", "pct_single_mother", "pct_foreign_households",
                    "mean_building_age", "median_building_age", "pct_pre1980", "total_dwellings")

predictor_vars <- c("mean_equiv_income", "total_population", "pct_foreign", 
                    "pct_single_person", "pct_foreign_pop", "total_households_register", 
                    "n_single_parent", "n_single_mother")

model_vars <- unique(c(vars_to_impute, predictor_vars))
model_vars <- model_vars[model_vars %in% names(census_ebp)]

cat("Sections for CV:", nrow(census_ebp), "\n")
cat("Model variables:", length(model_vars), "\n")
```

##### CV Function

```{r}
compare_methods_cv <- function(data, model_vars, vars_to_evaluate, 
                               n_reps = 10, pct_mask = 0.20, seed = 123) {
  
  complete_data <- data |> 
    select(all_of(c(model_vars, "income_quintile"))) |> 
    filter(complete.cases(pick(all_of(vars_to_evaluate))))
  
  results <- list()
  set.seed(seed)
  
  for(rep in 1:n_reps) {
    cat(sprintf("Rep %d/%d... ", rep, n_reps))
    
    data_rep <- complete_data
    true_values <- list()
    
    # Mask values
    for(var in vars_to_evaluate) {
      n_mask <- round(nrow(data_rep) * pct_mask)
      idx_mask <- sample(1:nrow(data_rep), n_mask)
      true_values[[var]] <- data_rep[[var]][idx_mask]
      data_rep[[var]][idx_mask] <- NA
    }
    
    # Method 1: missForest
    tryCatch({
      mat_mf <- data_rep |> select(all_of(model_vars)) |> as.data.frame()
      mf_result <- missForest(mat_mf, maxiter = 5, ntree = 50, verbose = FALSE)
      
      for(var in vars_to_evaluate) {
        idx_mask <- which(is.na(data_rep[[var]]))
        imputed_values <- mf_result$ximp[[var]][idx_mask]
        rmse <- sqrt(mean((true_values[[var]] - imputed_values)^2))
        results[[length(results) + 1]] <- data.frame(
          repetition = rep, method = "missForest", variable = var, rmse = rmse)
      }
    }, error = function(e) NULL)
    
    # Method 2: MICE
    tryCatch({
      mat_mice <- data_rep |> select(all_of(model_vars)) |> as.data.frame()
      mice_result <- mice(mat_mice, m = 1, maxit = 5, method = "pmm", 
                          printFlag = FALSE, seed = seed + rep)
      mice_complete <- complete(mice_result, 1)
      
      for(var in vars_to_evaluate) {
        idx_mask <- which(is.na(data_rep[[var]]))
        imputed_values <- mice_complete[[var]][idx_mask]
        rmse <- sqrt(mean((true_values[[var]] - imputed_values)^2))
        results[[length(results) + 1]] <- data.frame(
          repetition = rep, method = "MICE", variable = var, rmse = rmse)
      }
    }, error = function(e) NULL)
    
    # Method 3: Amelia
    tryCatch({
      mat_amelia <- data_rep |> select(all_of(model_vars)) |> as.data.frame()
      amelia_result <- amelia(mat_amelia, m = 1, p2s = 0, boot.type = "none")
      amelia_complete <- amelia_result$imputations[[1]]
      
      for(var in vars_to_evaluate) {
        idx_mask <- which(is.na(data_rep[[var]]))
        imputed_values <- amelia_complete[[var]][idx_mask]
        rmse <- sqrt(mean((true_values[[var]] - imputed_values)^2))
        results[[length(results) + 1]] <- data.frame(
          repetition = rep, method = "Amelia", variable = var, rmse = rmse)
      }
    }, error = function(e) NULL)
    
    # Method 4: Mean by quintile
    for(var in vars_to_evaluate) {
      idx_mask <- which(is.na(data_rep[[var]]))
      quintile_means <- data_rep |> 
        filter(!is.na(.data[[var]])) |> 
        group_by(income_quintile) |> 
        summarise(mean_value = mean(.data[[var]], na.rm = TRUE), .groups = "drop")
      
      imputed_values <- data_rep$income_quintile[idx_mask] |> 
        map_dbl(~quintile_means$mean_value[quintile_means$income_quintile == .x])
      
      rmse <- sqrt(mean((true_values[[var]] - imputed_values)^2))
      results[[length(results) + 1]] <- data.frame(
        repetition = rep, method = "Mean_Quintile", variable = var, rmse = rmse)
    }
    
    # Method 5: Median by quintile
    for(var in vars_to_evaluate) {
      idx_mask <- which(is.na(data_rep[[var]]))
      quintile_medians <- data_rep |> 
        filter(!is.na(.data[[var]])) |> 
        group_by(income_quintile) |> 
        summarise(median_value = median(.data[[var]], na.rm = TRUE), .groups = "drop")
      
      imputed_values <- data_rep$income_quintile[idx_mask] |> 
        map_dbl(~quintile_medians$median_value[quintile_medians$income_quintile == .x])
      
      rmse <- sqrt(mean((true_values[[var]] - imputed_values)^2))
      results[[length(results) + 1]] <- data.frame(
        repetition = rep, method = "Median_Quintile", variable = var, rmse = rmse)
    }
  }
  
  bind_rows(results)
}
```

##### Run CV

```{r}
vars_to_evaluate <- c("pct_below_threshold_60", "pct_below_threshold_40", 
                      "net_household_income", "pct_over_65")

cv_results <- compare_methods_cv(
  data = census_ebp,
  model_vars = model_vars,
  vars_to_evaluate = vars_to_evaluate,
  n_reps = 10,
  pct_mask = 0.20,
  seed = 123
)
```

#### Results

```{r}
global_summary <- cv_results |> 
  group_by(method) |> 
  summarise(mean_rmse = mean(rmse), sd_rmse = sd(rmse), .groups = "drop") |> 
  arrange(mean_rmse)

print(global_summary)
```

##### Visualization

```{r}
# Select sections with complete data for one key variable
demo_var <- "pct_below_threshold_60"

demo_data <- census_ebp |> 
  select(all_of(c(model_vars, "income_quintile"))) |> 
  filter(complete.cases(pick(all_of(demo_var))))

# Create a copy with 20% masked
set.seed(123)
n_mask <- round(nrow(demo_data) * 0.20)
idx_mask <- sample(1:nrow(demo_data), n_mask)

demo_masked <- demo_data
demo_masked[[demo_var]][idx_mask] <- NA

# Store original values
original_values <- demo_data[[demo_var]][idx_mask]

# Impute with each method
imputed_values <- data.frame(original = original_values)

# missForest
mat_mf <- demo_masked |> select(all_of(model_vars)) |> as.data.frame()
mf_result <- missForest(mat_mf, maxiter = 5, ntree = 50, verbose = FALSE)
imputed_values$missForest <- mf_result$ximp[[demo_var]][idx_mask]

# MICE
mat_mice <- demo_masked |> select(all_of(model_vars)) |> as.data.frame()
mice_result <- mice(mat_mice, m = 1, maxit = 5, method = "pmm", printFlag = FALSE, seed = 123)
mice_complete <- complete(mice_result, 1)
imputed_values$MICE <- mice_complete[[demo_var]][idx_mask]

# Amelia
mat_amelia <- demo_masked |> select(all_of(model_vars)) |> as.data.frame()
amelia_result <- amelia(mat_amelia, m = 1, p2s = 0, boot.type = "none")
amelia_complete <- amelia_result$imputations[[1]]
imputed_values$Amelia <- amelia_complete[[demo_var]][idx_mask]

# Mean by quintile
quintile_means <- demo_masked |> 
  filter(!is.na(.data[[demo_var]])) |> 
  group_by(income_quintile) |> 
  summarise(mean_value = mean(.data[[demo_var]], na.rm = TRUE), .groups = "drop")
imputed_values$Mean_Quintile <- demo_masked$income_quintile[idx_mask] |> 
  map_dbl(~quintile_means$mean_value[quintile_means$income_quintile == .x])

# Pivot to long format
variables <- c("original", "missForest", "MICE", "Amelia", "Mean_Quintile")
titles <- c("Original Distribution", "missForest", "MICE (PMM)", "Amelia (EM)", "Mean by Quintile")
colors_fill <- c("skyblue", "#15ad4f", "#6a6ad9", "#e65100", "#ff6f00")

imputed_long <- imputed_values |> 
  pivot_longer(all_of(variables), names_to = "method", values_to = "value")

# Create faceted histogram
fig_b1 <- imputed_long |> 
  mutate(title = factor(method, levels = variables, labels = titles)) |> 
  ggplot(aes(x = value, fill = title)) +
  geom_histogram(binwidth = 2, color = "#808080", alpha = 0.8) +
  facet_wrap(~title, scales = "free_y", ncol = 3) +
  scale_fill_manual(values = colors_fill) +
  labs(title = "Distribution Comparison: Original vs Imputed Values",
       subtitle = paste0("Variable: ", demo_var, " (", n_mask, " masked values)"),
       x = "Value", y = "Frequency") +
  theme_classic() +
  theme(legend.position = "none",
        strip.text = element_text(face = "bold", size = 11),
        plot.title = element_text(size = 14, face = "bold"))

fig_b1
```

##### Error Distribution Analysis

```{r}
# Calculate errors for each method
error_data <- data.frame(
  missForest = imputed_values$missForest - imputed_values$original,
  MICE = imputed_values$MICE - imputed_values$original,
  Amelia = imputed_values$Amelia - imputed_values$original,
  Mean_Quintile = imputed_values$Mean_Quintile - imputed_values$original
)

# Pivot to long format
error_long <- error_data |> 
  pivot_longer(everything(), names_to = "method", values_to = "error")

# Summary statistics
error_summary <- error_long |> 
  group_by(method) |> 
  summarise(
    mean_error = mean(error),
    median_error = median(error),
    mad_error = mean(abs(error)),  # Mean Absolute Deviation
    rmse = sqrt(mean(error^2)),
    .groups = "drop"
  ) |> 
  arrange(rmse)

print(error_summary)

# Plot 1: Error distribution (histogram)
p1_err_dist <- ggplot(error_long, aes(x = error, fill = method)) +
  geom_histogram(bins = 30, alpha = 0.7, position = "identity") +
  geom_vline(xintercept = 0, linetype = "dashed", color = "red", size = 1) +
  facet_wrap(~method, ncol = 2) +
  scale_fill_manual(values = c("missForest" = "#15ad4f", "MICE" = "#6a6ad9", 
                                "Amelia" = "#e65100", "Mean_Quintile" = "#ff6f00")) +
  labs(title = "Error Distribution by Method",
       subtitle = "Error = Imputed - Original (0 = perfect prediction)",
       x = "Error (percentage points)", y = "Frequency") +
  theme_classic() +
  theme(legend.position = "none",
        strip.text = element_text(face = "bold", size = 11))

# Plot 2: Absolute error boxplot
p2_abe <- error_long |> 
  mutate(abs_error = abs(error)) |> 
  ggplot(aes(x = reorder(method, abs_error, median), y = abs_error, fill = method)) +
  geom_boxplot(alpha = 0.7) +
  geom_jitter(width = 0.2, alpha = 0.3, size = 0.5) +
  scale_fill_manual(values = c("missForest" = "#15ad4f", "MICE" = "#6a6ad9", 
                                "Amelia" = "#e65100", "Mean_Quintile" = "#ff6f00")) +
  labs(title = "Absolute Error Comparison",
       subtitle = "Lower is better",
       x = "Method", y = "Absolute Error (pp)") +
  theme_classic() +
  theme(legend.position = "none",
        axis.text.x = element_text(angle = 45, hjust = 1))

# Plot 3: Error vs Original value (scatterplot)
p3_vs <- bind_cols(
  original = rep(imputed_values$original, 4),
  error = error_long$error,
  method = error_long$method
) |> 
  ggplot(aes(x = original, y = error, color = method)) +
  geom_point(alpha = 0.5, size = 2) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
  geom_smooth(method = "loess", se = FALSE, size = 1) +
  facet_wrap(~method, ncol = 2) +
  scale_color_manual(values = c("missForest" = "#15ad4f", "MICE" = "#6a6ad9", 
                                 "Amelia" = "#e65100", "Mean_Quintile" = "#ff6f00")) +
  labs(title = "Error Pattern by Original Value",
       subtitle = "Do methods systematically over/under-estimate?",
       x = "Original Value", y = "Error (Imputed - Original)") +
  theme_classic() +
  theme(legend.position = "none",
        strip.text = element_text(face = "bold"))

# Combine plots
error_analysis <- (p1_err_dist | p2_abe) / p3_vs +
  plot_annotation(
    title = "Comprehensive Error Analysis",
    subtitle = paste0("Variable: ", demo_var, " | n = ", n_mask, " imputed values"),
    theme = theme(plot.title = element_text(size = 16, face = "bold"))
  )

error_analysis
```

##### Mean Absolute Error Ranking

```{r}
# Bar plot of MAE
mae_method <- ggplot(error_summary, aes(x = reorder(method, mad_error), y = mad_error, fill = method)) +
  geom_col(alpha = 0.8) +
  geom_text(aes(label = sprintf("%.2f pp", mad_error)), vjust = -0.5, size = 5) +
  scale_fill_manual(values = c("missForest" = "#15ad4f", "MICE" = "#6a6ad9", 
                                "Amelia" = "#e65100", "Mean_Quintile" = "#ff6f00")) +
  labs(title = "Mean Absolute Error (MAE) by Method",
       subtitle = "Average prediction error in percentage points",
       x = "Method", y = "MAE (percentage points)") +
  theme_classic() +
  theme(legend.position = "none",
        axis.text.x = element_text(angle = 45, hjust = 1, size = 12),
        plot.title = element_text(size = 14, face = "bold"))

mae_method
```

#### Imputation

Based on comprehensive cross-validation analysis comparing five imputation methods 
across four critical variables, **missForest** was selected as the optimal approach.

**Selection criteria:**

| Metric | missForest | MICE | Amelia | Mean Quintile |
|--------|-----------|------|--------|---------------|
| **Mean Absolute Error** | **0.95 pp** | 1.32 pp | 1.33 pp | 3.87 pp |
| **RMSE** | **5.87** | 5.99 | 5.98 | 5.70 |
| **Error distribution** | Centered at 0 | Good | Good | Biased |
| **Systematic bias** | None detected | Minimal | Moderate | High |

**Key advantages of missForest:**

1. **Lowest prediction error**: 28% lower MAE than MICE (0.95 vs 1.32 pp)
2. **Unbiased**: Error distribution centered at 0 with minimal variance
3. **Robust**: No systematic over/under-estimation across value ranges
4. **Non-linear modeling**: Captures complex relationships between predictors
5. **Standard method**: Widely used in Small Area Estimation literature

**Missing mechanism:** Data are **Missing at Random (MAR)** - missingness is 
related to observable characteristics (small population size) but not to the 
missing values themselves.

**Algorithm parameters:**
- Method: Random Forest multivariate imputation
- `maxiter = 10`: Maximum iterations for convergence
- `ntree = 100`: Trees per Random Forest
- `seed = 123`: Reproducibility
- Quality target: OOB error < 0.05 (excellent) or < 0.10 (acceptable)

```{r}
# Note: census_ebp and model_vars already defined in CV setup above
# If running independently, ensure they exist or uncomment below:
# census_ebp <- master_census |> 
#   filter(!is.na(mean_equiv_income)) |> 
#   mutate(income_quintile = ntile(mean_equiv_income, 5))

# Show variables with NAs
for(var in model_vars) {
  n_na <- sum(is.na(census_ebp[[var]]))
  if(n_na > 0) {
    cat(sprintf("  %-30s: %4d NAs (%5.1f%%)\n", var, n_na, 
                100 * n_na / nrow(census_ebp)))
  }
}

# Run missForest
set.seed(123)
mat_mf <- census_ebp |> select(all_of(model_vars)) |> as.data.frame()
total_nas <- sum(is.na(mat_mf))
cat(sprintf("Total NAs to impute: %d\n", total_nas))

start_time <- Sys.time()
mf_result <- missForest(mat_mf, maxiter = 10, ntree = 100, verbose = TRUE)
elapsed_time <- as.numeric(difftime(Sys.time(), start_time, units = "mins"))

if(mf_result$OOBerror < 0.05) {
  cat("(Excellent)\n")
} else if(mf_result$OOBerror < 0.10) {
  cat("(Good)\n")
} else {
  cat("(Acceptable)\n")
}

# Integrate imputed values
for(var in model_vars) {
  was_na <- is.na(census_ebp[[var]])
  n_imputed <- sum(was_na)
  
  if(n_imputed > 0) {
    census_ebp[[var]] <- mf_result$ximp[[var]]
    census_ebp[[paste0("imputed_", var)]] <- ifelse(was_na, "missForest", "observed")
    cat(sprintf("  %-30s: %4d imputed\n", var, n_imputed))
  }
}

# Verify completion
remaining_nas <- census_ebp |> 
  select(all_of(model_vars)) |> 
  summarise(across(everything(), ~sum(is.na(.)))) |> 
  pivot_longer(everything()) |> 
  filter(value > 0)

if(nrow(remaining_nas) == 0) {
  cat("\nAll NAs successfully imputed\n")
} else {
  cat("\nRemaining NAs:\n")
  print(remaining_nas)
}
```

#### Validation

```{r}
check_var <- "pct_below_threshold_60"

comparison <- master_census |> 
  select(CUSEC, original = all_of(check_var)) |> 
  inner_join(
    census_ebp |> select(CUSEC, imputed = all_of(check_var)),
    by = "CUSEC"
  ) |> 
  mutate(was_na = is.na(original))

p1_1 <- ggplot(comparison |> filter(!was_na), aes(x = original)) +
  geom_histogram(bins = 30, fill = "skyblue", color = "#808080") +
  labs(title = "A) Observed Values", x = check_var, y = "Frequency") +
  theme_classic()

p2_1 <- ggplot(comparison |> filter(was_na), aes(x = imputed)) +
  geom_histogram(bins = 30, fill = "#15ad4f", color = "#808080") +
  labs(title = "B) Imputed Values", x = check_var, y = "Frequency") +
  theme_classic()

p3_1 <- ggplot(comparison, aes(x = imputed, fill = was_na)) +
  geom_histogram(bins = 30, alpha = 0.7, position = "identity") +
  scale_fill_manual(values = c("FALSE" = "skyblue", "TRUE" = "#15ad4f"),
                    labels = c("Observed", "Imputed"), name = "") +
  labs(title = "C) Combined Distribution", x = check_var, y = "Frequency") +
  theme_classic() +
  theme(legend.position = "bottom")

imputation_quality <- (p1_1 | p2_1) / p3_1 +
  plot_annotation(
    title = "missForest Imputation Quality Check",
    subtitle = paste0("Variable: ", check_var, " (", sum(comparison$was_na), " imputed)")
  )

imputation_quality
```

```{r}
cat("Dimensions:", nrow(census_ebp), "sections ×", ncol(census_ebp), "variables\n")
```

## Phase 2: Small area estimation

### Overview

The ECV survey does not publish census section identifiers (CUSEC) in its public
microdata, as its sampling design is only representative at the regional level. 
This prevents direct estimation of energy poverty at the section level, requiring
the use of Small Area Estimation (SAE) techniques.
We implement a synthetic regression estimator (Rao & Molina, 2015, Section 3.3)
using a generalized linear mixed model with binomial family and logit link, appropriate
for the binary nature of the dependent variable. The model is trained on
household-level survey data (ECV Madrid, ~1,900 households) and applied to
section-level aggregated census covariates (Atlas INE, 2,443 sections) to
generate energy poverty predictions.

The model specification includes equivalized household income, household size,
and presence of elderly members (65+) as fixed effects, with random intercepts
by income quintile to account for unobserved heterogeneity across socioeconomic
strata. Since census sections are not sampled in the ECV, section-specific random
effects cannot be estimated and are assumed to equal their expected value of zero
(synthetic assumption).

Income quintile thresholds are calculated exclusively from the ECV distribution
and subsequently applied to classify census sections, ensuring methodological
consistency. The model is validated through correlations with independent Atlas
INE indicators: poverty rate (r = 0.92) and mean income (r = -0.90), confirming
that predicted energy vulnerability captures the socioeconomic structure of
census sections despite the micro-macro aggregation level difference.
This approach follows the World Bank guidelines for poverty mapping when
geographic identifiers are unavailable in survey data (Corral et al., 2022).
Complete theoretical framework, model diagnostics, and methodological
justification are provided in the thesis document.

### Step 1: Charge and prepare data

```{r}
# Load survey (Phase 1 output)
survey_raw <- read_csv("data/datos_fase1_final.csv", show_col_types = FALSE)

# Load imputed census (Phase 2 output - Imputation)
census_raw <- census_ebp

cat("Survey:", nrow(survey_raw), "households | Census:", nrow(census_raw), "sections\n\n")

valid_survey <- survey_raw |> 
  filter(!is.na(pobreza_energetica)) |>  
  mutate(
    Y = pobreza_energetica,
    
    income_raw = pmin(pmax(renta_equiv_euros, 7450), 51860),
    
    # Standardize
    income = income_raw / 10000,
    
    household_size = as.numeric(n_miembros),
    has_elderly_num = as.numeric(tiene_mayor65)
  ) |> 
  filter(!is.na(Y), !is.na(income))

cat("Valid households:", nrow(valid_survey), 
    "(excluded:", nrow(survey_raw) - nrow(valid_survey), ")\n")
cat("Mean energy poverty:", round(mean(valid_survey$Y) * 100, 1), "%\n\n")
```

### Step 2: Calculate quintiles from ECV data

```{r}
# CRITICAL: Quintile thresholds calculated ONLY from ECV
quintile_thresholds <- quantile(valid_survey$income, c(0.20, 0.40, 0.60, 0.80))

cat("Quintile thresholds (€/10k):\n")
cat(sprintf("  Q1: income ≤ %.3f (≤€%s)\n", quintile_thresholds[1], 
            format(quintile_thresholds[1] * 10000, big.mark = ",")))
cat(sprintf("  Q2: %.3f < income ≤ %.3f (€%s - €%s)\n", 
            quintile_thresholds[1], quintile_thresholds[2],
            format(quintile_thresholds[1] * 10000, big.mark = ","),
            format(quintile_thresholds[2] * 10000, big.mark = ",")))
cat(sprintf("  Q3: %.3f < income ≤ %.3f (€%s - €%s)\n", 
            quintile_thresholds[2], quintile_thresholds[3],
            format(quintile_thresholds[2] * 10000, big.mark = ","),
            format(quintile_thresholds[3] * 10000, big.mark = ",")))
cat(sprintf("  Q4: %.3f < income ≤ %.3f (€%s - €%s)\n", 
            quintile_thresholds[3], quintile_thresholds[4],
            format(quintile_thresholds[3] * 10000, big.mark = ","),
            format(quintile_thresholds[4] * 10000, big.mark = ",")))
cat(sprintf("  Q5: income > %.3f (>€%s)\n\n", quintile_thresholds[4],
            format(quintile_thresholds[4] * 10000, big.mark = ",")))

# Apply THESE SAME thresholds to the ECV
valid_survey <- valid_survey |> 
  mutate(
    household_quintile = factor(case_when(
      income <= quintile_thresholds[1] ~ 1,
      income <= quintile_thresholds[2] ~ 2,
      income <= quintile_thresholds[3] ~ 3,
      income <= quintile_thresholds[4] ~ 4,
      TRUE ~ 5
    ), levels = 1:5)
  )
```

### Step 3: apply previous quintiles to censo data

```{r}
valid_census <- census_raw |> 
  mutate(
    # Standardize census income
    income = mean_equiv_income / 10000,
    
    household_size = mean_household_size,
    has_elderly_num = pct_over_65 / 100,
    
    # CRITICAL: Apply SAME thresholds from ECV
    section_quintile = case_when(
      income <= quintile_thresholds[1] ~ 1,
      income <= quintile_thresholds[2] ~ 2,
      income <= quintile_thresholds[3] ~ 3,
      income <= quintile_thresholds[4] ~ 4,
      TRUE ~ 5
    ),
    
    # For the model, call it household_quintile (same name as in ECV)
    household_quintile = factor(section_quintile, levels = 1:5)
  ) |> 
  filter(!is.na(income), !is.na(household_quintile)) |> 
  select(CUSEC, section_quintile, household_quintile, income, household_size, 
         has_elderly_num, pct_below_threshold_60, 
         gini, pct_pre1980, total_population, median_equiv_income)

# Verify distribution in Atlas (will NOT be uniform - this is correct)
cat("Atlas distribution by quintile (using ECV thresholds):\n")
census_dist <- valid_census |> 
  count(section_quintile) |> 
  mutate(pct = round(100 * n / sum(n), 1))
print(census_dist)
```

### Step 4: estimate mixed logitis model (training phase)

```{r}
sae_model <- glmer(
  Y ~ income + household_size + has_elderly_num + (1 | household_quintile),
  data = valid_survey,
  family = binomial(link = "logit"),
  control = glmerControl(
    optimizer = "bobyqa", 
    optCtrl = list(maxfun = 100000)
  )
)

# =============================================================================
# Model summary
# =============================================================================

cat("FIXED EFFECTS (log-odds scale):\n")
coef_summary <- summary(sae_model)$coefficients
coef_df <- as.data.frame(coef_summary) |> 
  rownames_to_column("Coefficient") |> 
  mutate(
    OR = exp(Estimate),
    sig = case_when(
      `Pr(>|z|)` < 0.001 ~ "***",
      `Pr(>|z|)` < 0.01 ~ "**",
      `Pr(>|z|)` < 0.05 ~ "*",
      TRUE ~ ""
    )
  ) |> 
  select(Coefficient, Estimate, `Std. Error`, OR, `z value`, `Pr(>|z|)`, sig)

print(coef_df, digits = 4)

cat("\nInterpretation (Odds Ratios):\n")
cat(sprintf("  income:             OR = %.3f (%.1f%% decrease per €10k)\n", 
            exp(coef(summary(sae_model))["income", "Estimate"]),
            (1 - exp(coef(summary(sae_model))["income", "Estimate"])) * 100))
cat(sprintf("  household_size:     OR = %.3f (%.1f%% increase per member)\n",
            exp(coef(summary(sae_model))["household_size", "Estimate"]),
            (exp(coef(summary(sae_model))["household_size", "Estimate"]) - 1) * 100))
cat(sprintf("  has_elderly_num:    OR = %.3f\n",
            exp(coef(summary(sae_model))["has_elderly_num", "Estimate"])))

cat("RANDOM EFFECTS:\n")
print(VarCorr(sae_model))

cat("\nRandom effects by quintile:\n")
ranef_vals <- ranef(sae_model)$household_quintile
ranef_df <- data.frame(
  Quintile = 1:5,
  Effect = ranef_vals[,1]
) |> 
  mutate(
    Direction = ifelse(Effect > 0, "+", "-"),
    Interpretation = case_when(
      Effect > 0.05 ~ "Higher poverty than expected",
      Effect < -0.05 ~ "Lower poverty than expected",
      TRUE ~ "Near average"
    )
  )
print(ranef_df, digits = 4)

cat("\n")
cat("MODEL FIT:\n")
cat(sprintf("  AIC:  %.1f\n", AIC(sae_model)))
cat(sprintf("  BIC:  %.1f\n", BIC(sae_model)))
cat(sprintf("  Log-Likelihood: %.1f\n", logLik(sae_model)))

# R²
r2 <- r.squaredGLMM(sae_model)
cat(sprintf("  R² marginal:    %.3f (fixed effects only)\n", r2[1]))
cat(sprintf("  R² conditional: %.3f (fixed + random effects)\n", r2[2]))
```

### Step 5: synthetic prediction for non-sampled areas

```{r}
cat("Method: Synthetic Regression Estimator (Rao & Molina, 2015, Section 3.3)\n")
cat("Assumption: v_section = E[v] = 0 (no data available to estimate section-specific effects)\n\n")

cat("Prediction formula for section j:\n")
cat("  logit(p_j) = β̂₀ + β̂₁·income_j + β̂₂·household_size_j + β̂₃·elderly_j + v̂_i\n")
cat("  where v̂_i = random effect of quintile i (from training)\n\n")

# Prediction WITH random effects (uses v̂_i from quintile)
cat("Generating predictions with quintile random effects...\n")

evi_pred_with_re <- predict(
  sae_model, 
  newdata = valid_census,
  allow.new.levels = FALSE,
  re.form = NULL,
  type = "response"
)

# Prediction WITHOUT random effects (v = 0)
evi_pred_without_re <- predict(
  sae_model,
  newdata = valid_census,
  allow.new.levels = FALSE,
  re.form = ~0,
  type = "response"
)

# Integrate both predictions
sae_results <- valid_census |> 
  mutate(
    EVI_synthetic = evi_pred_with_re * 100,
    EVI_fixed_only = evi_pred_without_re * 100
  )

cat(sprintf("\nPredictions generated for %d sections\n", nrow(sae_results)))
cat("\nMETHOD 1: Synthetic with quintile effects (RECOMMENDED):\n")
cat(sprintf("  Mean EVI = %.1f%% | Range: [%.1f%%, %.1f%%]\n",
            mean(sae_results$EVI_synthetic),
            min(sae_results$EVI_synthetic),
            max(sae_results$EVI_synthetic)))

cat("\nMETHOD 2: Fixed effects only (no random effects):\n")
cat(sprintf("  Mean EVI = %.1f%% | Range: [%.1f%%, %.1f%%]\n\n",
            mean(sae_results$EVI_fixed_only),
            min(sae_results$EVI_fixed_only),
            max(sae_results$EVI_fixed_only)))

# Detailed statistics
cat("EVI SUMMARY (Synthetic with quintile effects):\n")
summary_stats <- summary(sae_results$EVI_synthetic)
print(summary_stats)

cat("\n\nDISTRIBUTION BY QUINTILE:\n")
quintile_dist <- sae_results |> 
  group_by(section_quintile) |> 
  summarise(
    n_sections = n(),
    pct_sections = round(100 * n() / nrow(sae_results), 1),
    EVI_mean = mean(EVI_synthetic),
    EVI_sd = sd(EVI_synthetic),
    EVI_min = min(EVI_synthetic),
    EVI_max = max(EVI_synthetic),
    .groups = "drop"
  ) |> 
  mutate(across(where(is.numeric) & !c(n_sections, pct_sections), ~round(., 1)))

print(quintile_dist)
```

### Step 6: external validation

```{r}
correlations <- sae_results |> 
  summarise(
    cor_income = cor(EVI_synthetic / 100, income, use = "complete.obs"),
    cor_threshold60 = cor(EVI_synthetic / 100, pct_below_threshold_60, use = "complete.obs"),
    cor_gini = cor(EVI_synthetic / 100, gini, use = "complete.obs"),
    cor_pre1980 = cor(EVI_synthetic / 100, pct_pre1980, use = "complete.obs")
  )

cat("Correlations between predicted EVI and Atlas INE indicators:\n\n")

if (!is.na(correlations$cor_income)) {
  cat(sprintf("  %-30s r = %+.3f", "Mean income:", correlations$cor_income))
  if (correlations$cor_income < -0.7) {
    cat("  Strong negative (expected)\n")
  } else if (correlations$cor_income < 0) {
    cat("  Negative (expected)\n")
  } else {
    cat("  Unexpected sign\n")
  }
}

if (!is.na(correlations$cor_threshold60)) {
  cat(sprintf("  %-30s r = %+.3f", "Poverty rate (60% threshold):", correlations$cor_threshold60))
  if (correlations$cor_threshold60 > 0.7) {
    cat("  Strong positive (expected)\n")
  } else if (correlations$cor_threshold60 > 0) {
    cat("  Positive (expected)\n")
  } else {
    cat("  Unexpected sign\n")
  }
}

if (!is.na(correlations$cor_gini)) {
  cat(sprintf("  %-30s r = %+.3f", "Gini coefficient:", correlations$cor_gini))
  cat("  (informative)\n")
}

if (!is.na(correlations$cor_pre1980)) {
  cat(sprintf("  %-30s r = %+.3f", "%% dwellings pre-1980:", correlations$cor_pre1980))
  if (correlations$cor_pre1980 > 0) {
    cat("   Positive (older = less efficient)\n")
  } else {
    cat("  ~ (informative)\n")
  }
}
```

### Step 7: visualizations

```{r}
# Plot 1: EVI Distribution
p_dist <- ggplot(sae_results, aes(x = EVI_synthetic)) +
  geom_histogram(bins = 50, fill = "#2A9D8F", color = "white", alpha = 0.8) +
  geom_vline(xintercept = mean(sae_results$EVI_synthetic),
             color = "#E63946", linewidth = 1, linetype = "dashed") +
  labs(
    title = "Energy Vulnerability Index Distribution",
    subtitle = "Synthetic estimation for 2,443 census sections in Madrid",
    x = "EVI (%)",
    y = "Number of sections",
    caption = paste0("Mean = ", round(mean(sae_results$EVI_synthetic), 1), "%")
  ) +
theme_minimal(base_size = 10) +    
theme(plot.title = element_text(face = "bold", size = 10),  
      plot.subtitle = element_text(size = 8))                 

# Plot 2: EVI vs Income
p_income <- ggplot(sae_results, aes(x = income * 10000, y = EVI_synthetic)) +
  geom_point(alpha = 0.4, color = "#457B9D", size = 1) +
  geom_smooth(method = "loess", color = "#E63946", se = TRUE, linewidth = 1.2) +
  scale_x_continuous(labels = scales::dollar_format(prefix = "€", big.mark = ",")) +
  labs(
    title = "EVI vs Mean Income per Equivalence Unit",
    subtitle = paste0("Correlation: r = ", round(correlations$cor_income, 3)),
    x = "Mean Income (€/EU)",
    y = "EVI (%)"
  ) +
theme_minimal(base_size = 10) +    
theme(plot.title = element_text(face = "bold", size = 10),  
      plot.subtitle = element_text(size = 8)) 

# Plot 3: EVI by Quintile
p_quint <- ggplot(sae_results, aes(x = factor(section_quintile), y = EVI_synthetic)) +
  geom_boxplot(fill = "#2A9D8F", alpha = 0.7, outlier.alpha = 0.3) +
  stat_summary(fun = mean, geom = "point", shape = 23, size = 3, 
               fill = "#E63946", color = "white") +
  labs(
    title = "EVI by Income Quintile",
    subtitle = "Red diamond = mean | Box = IQR",
    x = "Income Quintile (1=lowest, 5=highest)",
    y = "EVI (%)"
  ) +
theme_minimal(base_size = 10) +    
theme(plot.title = element_text(face = "bold", size = 10),  
      plot.subtitle = element_text(size = 8)) 

# Plot 4: EVI vs % below 60% threshold
p_poverty <- ggplot(sae_results, aes(x = pct_below_threshold_60, y = EVI_synthetic)) +
  geom_point(alpha = 0.4, color = "#457B9D", size = 1) +
  geom_smooth(method = "lm", color = "#E63946", se = TRUE, linewidth = 1.2) +
  labs(
    title = "EVI vs Population Below Poverty Threshold (60%)",
    subtitle = paste0("Correlation: r = ", round(correlations$cor_threshold60, 3)),
    x = "% population below 60% threshold",
    y = "EVI (%)"
  ) +
theme_minimal(base_size = 10) +    
theme(plot.title = element_text(face = "bold", size = 10),  
      plot.subtitle = element_text(size = 8)) 

# Combine plots
combined_plot <- (p_dist | p_income) / (p_quint | p_poverty) +
  plot_annotation(
    title = "Small Area Estimation: Synthetic Regression Estimator (Logistic Mixed Model)",
    subtitle = "Method: Rao & Molina (2015, Section 3.2) | Application: Energy Poverty in Madrid",
    theme = theme(
      plot.title = element_text(size = 13, face = "bold"),
      plot.subtitle = element_text(size = 9)                 
    )
  )

combined_plot

fig_4_2 <- p_income | p_poverty +
  plot_annotation(
    title = "Figure 4.2 — External validation of the synthetic EVI",
    subtitle = "Bivariate relationships with two independent INE Atlas indicators (n = 2,443 sections)",
    theme = theme(
      plot.title = element_text(size = 15, face = "bold", hjust = 0.5),
      plot.subtitle = element_text(size = 11, color = "gray40", hjust = 0.5)
    )
  )
```

### Summary

```{r}
cat("Method: Synthetic Regression Estimator (Rao & Molina, 2015)\n")
cat("Application: Energy Poverty Index for census sections in Madrid\n")
cat("Training: ", nrow(valid_survey), " households (ECV)\n")
cat("Prediction: ", nrow(sae_results), " sections (Atlas INE)\n")
cat("Mean EVI: ", round(mean(sae_results$EVI_synthetic), 1), "%\n")
cat("Validation: r(poverty_60%) = ", round(correlations$cor_threshold60, 3), 
    " | r(income) = ", round(correlations$cor_income, 3), "\n\n")
```

## Umbral justification

### Threshold Justification

The Synthetic Regression Estimator from Phase 2 produces continuous predictions 
of the Energy Vulnerability Index (EVI) for each census section, but identifying 
sections as *energy poor* requires defining a cut-off point. Following the 
literature on energy poverty measurement, this study adopts **EVI > 15** as the 
classification threshold.

To ensure that this choice is methodologically robust and not arbitrary, the 
threshold is validated through three independent approaches:

1. **Coherence with the Eurostat criterion**: comparing the proposed threshold 
   against the European standard for identifying excessive spending on basic 
   needs (gasto > 2× median).

2. **Empirical distribution in Madrid**: assessing the position of the threshold 
   within the local EVI distribution and the resulting prevalence of energy 
   poverty against published estimates for Spain.

3. **Validation with independent indicators (Atlas INE)**: testing whether 
   sections classified as energy poor (EVI > 15) exhibit statistically 
   significant differences in socioeconomic indicators that were *not* used 
   in the SAE model (poverty rate, mean income, Gini coefficient).

The convergence of these three approaches provides empirical support for the 
chosen threshold and ensures that subsequent classification captures real 
socioeconomic vulnerability rather than methodological artifacts.

#### Eurostat Criterion

```{r}
# ------------------------------------------------------------------------------
# EUROSTAT CRITERION: Threshold as a multiple of the median
# ------------------------------------------------------------------------------

# Compute descriptive statistics
median_evi <- median(sae_results$EVI_synthetic, na.rm = TRUE)
q1_evi     <- quantile(sae_results$EVI_synthetic, 0.25, na.rm = TRUE)
q3_evi     <- quantile(sae_results$EVI_synthetic, 0.75, na.rm = TRUE)
mean_evi   <- mean(sae_results$EVI_synthetic, na.rm = TRUE)

# Threshold/median ratio
eurostat_ratio <- 15 / median_evi

cat(sprintf("  EVI median in Madrid:  %.2f\n", median_evi))
cat(sprintf("  Proposed threshold:    15.00\n"))
cat(sprintf("  Ratio (15/median):     %.2fx\n\n", eurostat_ratio))

if (eurostat_ratio >= 1.3 && eurostat_ratio <= 2.5) {
  cat("  COHERENT with Eurostat criterion:\n")
  cat("    The 15/median ratio falls within the European standard range (1.3x - 2.5x)\n")
  cat("    for identifying excessive spending on basic needs.\n")
  cat("    Eurostat criterion: spending > 2× median = excessive spending\n")
} else {
  cat(sprintf("  Ratio %.2fx is outside the typical Eurostat range (1.3-2.5x)\n", eurostat_ratio))
}
```
#### Empirical Distribution

```{r}
# Prevalence above the threshold
pct_above_15 <- 100 * mean(sae_results$EVI_synthetic > 15, na.rm = TRUE)
n_above_15   <- sum(sae_results$EVI_synthetic > 15, na.rm = TRUE)

cat(sprintf("  Mean:     %.2f\n", mean_evi))
cat(sprintf("  Median:   %.2f\n", median_evi))
cat(sprintf("  Q1 (p25): %.2f\n", q1_evi))
cat(sprintf("  Q3 (p75): %.2f\n", q3_evi))
cat("\n")
cat(sprintf("  Threshold 15 falls at the %.0fth percentile of the distribution\n", 
            100 * mean(sae_results$EVI_synthetic <= 15, na.rm = TRUE)))
cat(sprintf("  Sections with EVI > 15: %d of %d (%.1f%%)\n", 
            n_above_15, nrow(sae_results), pct_above_15))
cat("\n")

if (pct_above_15 >= 30 && pct_above_15 <= 50) {
  cat("  COHERENT with energy poverty estimates for Spain:\n")
  cat(sprintf("    - This estimate: %.1f%% of sections\n", pct_above_15))
  cat("    - Spanish literature: 25-45%% households (ACA 2021, Tirado 2018)\n")
  cat("    - European Energy Poverty Observatory: 30-40%% Spain\n")
} else if (pct_above_15 < 30) {
  cat(sprintf("  Estimate (%.1f%%) is conservative vs literature (30-45%%)\n", pct_above_15))
} else {
  cat(sprintf("  Estimate (%.1f%%) is higher than literature (30-45%%)\n", pct_above_15))
}
```

#### Validation with Atlas INE Indicators

```{r}
# Explicit join: Atlas INE + EVI_synthetic
validation_data <- census_ebp |> 
  select(CUSEC, pct_below_threshold_60, gini, pct_pre1980, mean_equiv_income) |> 
  inner_join(
    sae_results |> select(CUSEC, EVI_synthetic), 
    by = "CUSEC"
  ) |> 
  # Create binary variable: energy poor YES/NO
  mutate(poor_evi15 = EVI_synthetic > 15)

cat(sprintf("  Sections with complete data: %d\n", nrow(validation_data)))
cat(sprintf("  Poor sections (EVI>15):      %d (%.1f%%)\n", 
            sum(validation_data$poor_evi15), 
            100 * mean(validation_data$poor_evi15)))
cat("\n")

# Compare Atlas INE means between poor and non-poor sections
atlas_comparison <- validation_data |> 
  group_by(poor_evi15) |> 
  summarise(
    n                    = n(),
    mean_pct_threshold60 = mean(pct_below_threshold_60, na.rm = TRUE),
    mean_income          = mean(mean_equiv_income, na.rm = TRUE),
    mean_gini            = mean(gini, na.rm = TRUE),
    mean_pct_pre1980     = mean(pct_pre1980, na.rm = TRUE),
    .groups = "drop"
  )

cat("  Comparison: poor sections (EVI>15) vs non-poor (EVI≤15):\n\n")

cat(sprintf("  Variable                       Non-poor      Poor      Difference\n"))
cat(sprintf("  %-28s   %8.1f    %8.1f    %+8.1f%%%%\n",
            "%% below 60%% threshold (Atlas)",
            atlas_comparison$mean_pct_threshold60[1],
            atlas_comparison$mean_pct_threshold60[2],
            100 * (atlas_comparison$mean_pct_threshold60[2] / atlas_comparison$mean_pct_threshold60[1] - 1)))

cat(sprintf("  %-28s   %8.0f€   %8.0f€   %+8.1f%%%%\n",
            "Mean income (Atlas)",
            atlas_comparison$mean_income[1],
            atlas_comparison$mean_income[2],
            100 * (atlas_comparison$mean_income[2] / atlas_comparison$mean_income[1] - 1)))

cat(sprintf("  %-28s   %8.3f    %8.3f    %+8.1f%%%%\n",
            "Gini (Atlas)",
            atlas_comparison$mean_gini[1],
            atlas_comparison$mean_gini[2],
            100 * (atlas_comparison$mean_gini[2] / atlas_comparison$mean_gini[1] - 1)))

# Statistical tests
test_threshold60 <- t.test(
  pct_below_threshold_60 ~ poor_evi15,
  data = validation_data,
  alternative = "less"  # Poor sections expected to have MORE % below threshold
)

test_income <- t.test(
  mean_equiv_income ~ poor_evi15,
  data = validation_data,
  alternative = "greater"  # Poor sections expected to have LESS income
)

test_gini <- t.test(
  gini ~ poor_evi15,
  data = validation_data
)

cat("\n")
cat("  Statistical tests:\n")
cat(sprintf("    %% below 60%%%% threshold: t = %.2f, p-value = %.4f %s\n",
            test_threshold60$statistic,
            test_threshold60$p.value,
            ifelse(test_threshold60$p.value < 0.001, "***", 
                   ifelse(test_threshold60$p.value < 0.01, "**", 
                          ifelse(test_threshold60$p.value < 0.05, "*", "")))))
cat(sprintf("    Mean income:             t = %.2f, p-value = %.4f %s\n",
            test_income$statistic,
            test_income$p.value,
            ifelse(test_income$p.value < 0.001, "***", 
                   ifelse(test_income$p.value < 0.01, "**", 
                          ifelse(test_income$p.value < 0.05, "*", "")))))
cat(sprintf("    Gini:                    t = %.2f, p-value = %.4f %s\n",
            test_gini$statistic,
            test_gini$p.value,
            ifelse(test_gini$p.value < 0.001, "***", 
                   ifelse(test_gini$p.value < 0.01, "**", 
                          ifelse(test_gini$p.value < 0.05, "*", "")))))

cat("\n")
if (test_threshold60$p.value < 0.05 && test_income$p.value < 0.05) {
  cat("  SUCCESSFUL VALIDATION:\n")
  cat("    Sections with EVI>15 show significant differences in independent\n")
  cat("    socioeconomic indicators (Atlas INE), confirming that the threshold\n")
  cat("    captures real economic vulnerability.\n")
} else {
  cat("   Tests do not show significant differences.\n")
}
```

#### Visualizations

```{r}
# EVI Distribution with Threshold
p_hist <- sae_results |> 
  ggplot(aes(x = EVI_synthetic)) +
  geom_histogram(aes(fill = EVI_synthetic > 15), 
                 bins = 50, alpha = 0.85, color = "white") +
  geom_vline(xintercept = 15, linetype = "dashed", 
             color = "red", linewidth = 1.2) +
  geom_vline(xintercept = median_evi, linetype = "dotted", 
             color = "black", linewidth = 0.8) +
  annotate("text", x = 15.5, y = 140, 
           label = "Threshold = 15", 
           hjust = 0, color = "red", size = 4, fontface = "bold") +
  annotate("text", x = median_evi - 0.5, y = 120, 
           label = sprintf("Median = %.1f", median_evi), 
           hjust = 1, size = 3.5, color = "black") +
  scale_fill_manual(
    values = c("FALSE" = "#2E75B6", "TRUE" = "#C00000"),
    labels = c("FALSE" = "EVI ≤ 15  (non-poor)", 
               "TRUE" = "EVI > 15  (energy poor)")
  ) +
  labs(
    title = "A) EVI Distribution and Threshold",
    subtitle = sprintf("%.1f%% of sections classified as energy poor | Ratio 15/median = %.2fx",
                       pct_above_15, eurostat_ratio),
    x = "EVI (Energy Vulnerability Index)",
    y = "Number of census sections",
    fill = NULL
  ) +
  theme_classic(base_size = 11) +
  theme(
    plot.title = element_text(face = "bold", size = 13, hjust = 0),
    plot.subtitle = element_text(size = 10, color = "gray40", hjust = 0),
    legend.position = "top",
    legend.text = element_text(size = 10)
  )

# Validation with INE Atlas indicators (boxplots)
p_box_data <- validation_data |> 
  mutate(Group = ifelse(poor_evi15, 
                        "Energy poor\n(EVI > 15)", 
                        "Non-poor\n(EVI ≤ 15)")) |> 
  select(Group, pct_below_threshold_60, mean_equiv_income) |> 
  pivot_longer(cols = c(pct_below_threshold_60, mean_equiv_income),
               names_to = "Variable",
               values_to = "Value") |> 
  mutate(Variable = case_when(
    Variable == "pct_below_threshold_60" ~ "% population below 60% poverty threshold",
    Variable == "mean_equiv_income"      ~ "Mean equivalised income (€)"
  ))

p_box <- ggplot(p_box_data, aes(x = Group, y = Value, fill = Group)) +
  geom_boxplot(alpha = 0.85, outlier.alpha = 0.4, outlier.size = 1) +
  facet_wrap(~Variable, scales = "free_y") +
  scale_fill_manual(values = c("Non-poor\n(EVI ≤ 15)" = "#2E75B6", 
                               "Energy poor\n(EVI > 15)" = "#C00000")) +
  labs(
    title = "B) Validation against Independent Indicators (INE Atlas)",
    subtitle = "Energy-poor sections show significantly worse socioeconomic conditions (p < 0.001)",
    x = NULL, y = NULL
  ) +
  theme_classic(base_size = 11) +
  theme(
    plot.title = element_text(face = "bold", size = 13, hjust = 0),
    plot.subtitle = element_text(size = 10, color = "gray40", hjust = 0),
    legend.position = "none",
    strip.text = element_text(face = "bold", size = 10),
    strip.background = element_rect(fill = "gray95", color = NA),
    axis.text.x = element_text(size = 9.5)
  )


fig_c1 <- p_hist / p_box +
  plot_annotation(
    title = "Triple Validation of the EVI > 15 Threshold",
    subtitle = "Empirical distribution and comparison with independent INE Atlas indicators",
    theme = theme(
      plot.title = element_text(size = 15, face = "bold", hjust = 0.5),
      plot.subtitle = element_text(size = 11, color = "gray40", hjust = 0.5)
    )
  ) +
  plot_layout(heights = c(1.2, 1))


print(fig_c1)
```

#### Conclusion

```{r}
cat("1. EUROSTAT CRITERION:\n")
cat(sprintf("   Ratio 15/median = %.2fx\n", eurostat_ratio))
cat("   Coherent with European standards (1.3x - 2.5x) for excessive spending\n\n")

cat("2. EMPIRICAL DISTRIBUTION:\n")
cat(sprintf("   %.1f%% of sections classified as energy poor\n", pct_above_15))
cat("   Coherent with Spanish literature estimates (25-45%%)\n\n")

cat("3. ATLAS INE VALIDATION:\n")
cat("   Sections with EVI>15 show:\n")
cat(sprintf("     - %.0f%%%% MORE population below 60%%%% threshold (p<%.3f)\n",
            abs(100 * (atlas_comparison$mean_pct_threshold60[2] / atlas_comparison$mean_pct_threshold60[1] - 1)),
            test_threshold60$p.value))
cat(sprintf("     - %.0f%%%% LESS mean income (p<%.3f)\n",
            abs(100 * (atlas_comparison$mean_income[2] / atlas_comparison$mean_income[1] - 1)),
            test_income$p.value))
```
The triple validation confirms that the EVI > 15 threshold is statistically 
justified: it is consistent with European benchmarks (1.36× the median), 
corresponds to the 60th percentile of the Madrid distribution, and sections 
above this cutoff exhibit significantly worse socioeconomic conditions across 
all three independent INE Atlas indicators (p < 0.001).


## Phase 3: Random Forest Models

This phase builds two complementary Random Forest models that predict energy 
poverty at the census section level using Google Earth Engine (GEE) variables 
as the primary data source:

- **Model 1 — Energy Poverty Classifier**: predicts whether a section is 
  *energy poor* (EVI > 15) using GEE environmental variables combined with 
  building age data. This model evaluates whether satellite-derived features 
  can recover the socioeconomic classification produced by the SAE model in 
  Phase 2.

- **Model 2 — Physical Vulnerability Classifier**: predicts whether a section 
  presents adverse *physical* conditions (extreme temperatures, lack of urban 
  mitigation, inefficient buildings) using **only** GEE variables. This model 
  identifies sections with structural vulnerability that may not appear in 
  income-based surveys.

A discrepancy analysis then characterizes four groups of sections (concordant, 
non-vulnerable, spatial false positives, and spatial false negatives) to 
identify areas that are *invisible* to traditional survey-based methods but 
exhibit clear physical vulnerability signals from satellite data.

Note: A first version of Model 2 (V1) suffered from circularity — predictors 
were direct components of the target. This was corrected in V2 by excluding 
all variables used in the target construction from the predictor set.


### Step 0: prerequisities

```{r}
# Required objects from Phases 1 and 2
required_objects <- c("sae_results", "census_ebp")

missing_objects <- required_objects[!sapply(required_objects, exists)]

# Additional libraries needed (base ones already loaded in Phases 1/2)
if (!require(randomForest, quietly = TRUE)) install.packages("randomForest")
if (!require(caret, quietly = TRUE)) install.packages("caret")
if (!require(pROC, quietly = TRUE)) install.packages("pROC")

library(randomForest)
library(caret)
library(pROC)

# Configuration
set.seed(42)
options(scipen = 999)
```

### Step 1: Loading Google Earth Engine Data

```{r}
# Load GEE data
gee <- read_csv(gee_path, show_col_types = FALSE) |> 
  mutate(CUSEC = str_pad(as.character(CUSEC), 10, side = "left", pad = "0")) |> 
  select(
    CUSEC,
    starts_with("LST_winter"), starts_with("LST_summer"),
    starts_with("NDVI_"),
    starts_with("NDBI_"),
    starts_with("ALBEDO_"),
    tcc
  )

cat(sprintf("  %d sections with GEE data\n", nrow(gee)))
cat(sprintf("  %d GEE variables loaded\n", ncol(gee) - 1))
```

### Step 2: Integrating GEE with Phase 2 Data

```{r}
# Combine: GEE + EVI synthetic + Census
dataset <- gee |> 
  inner_join(
    sae_results |> select(CUSEC, EVI_synthetic, section_quintile),
    by = "CUSEC"
  ) |> 
  inner_join(
    census_ebp |> select(
      CUSEC, 
      mean_equiv_income,    # used in characterization + export
      mean_building_age,    # predictor M1 + characterization + export
      pct_pre1980           # predictor M1 + characterization + export
    ),
    by = "CUSEC"
  ) |> 
  drop_na()

cat(sprintf("  Integrated dataset: %d complete sections\n", nrow(dataset)))
cat(sprintf("  Total variables: %d\n", ncol(dataset)))
```

### Step 3: Feature Engineering (derived variables)

```{r}
dataset <- dataset |> 
  mutate(
    # ENERGY DEMAND
    heating_demand     = pmax(0, 18 - LST_winter_p10_21),
    cooling_demand     = pmax(0, LST_summer_p90_21 - 26),
    seasonal_amplitude = LST_summer_mean_21 - LST_winter_mean_21,
    
    # URBAN HEAT ISLAND INDEX
    # Combines: high built-up (NDBI) + high temperature (LST) + low vegetation (NDVI)
    urban_heat_island_raw = (NDBI_mean_21 * 0.4) + 
                            ((LST_summer_mean_21 / 40) * 0.4) - 
                            (NDVI_mean_21 * 0.2),
    urban_heat_island = (urban_heat_island_raw - min(urban_heat_island_raw, na.rm = TRUE)) / 
                        (max(urban_heat_island_raw, na.rm = TRUE) - 
                         min(urban_heat_island_raw, na.rm = TRUE))
  ) |> 
  select(-urban_heat_island_raw)
```

### Step 4: Building the Model 2 Target — Physical Vulnerability

```{r}
dataset <- dataset |> 
  mutate(
    # --------------------------------------------------------------------------
    # CRITERION 1: Extreme winter
    # --------------------------------------------------------------------------
    # Threshold: Coldest days below 25th percentile
    # Justification: Risk of condensation, humidity, high thermal losses
    extreme_winter = LST_winter_p10_21 < quantile(LST_winter_p10_21, 0.25, na.rm = TRUE),
    
    # --------------------------------------------------------------------------
    # CRITERION 2: Extreme summer
    # --------------------------------------------------------------------------
    # Threshold: Hottest days > 42°C (AEMET heat wave)
    # Justification: Indoor health risk without AC (>30°C indoor)
    extreme_summer = LST_summer_p90_21 > 42,
    
    # --------------------------------------------------------------------------
    # CRITERION 3: Inefficient buildings
    # --------------------------------------------------------------------------
    # Threshold: Majority of buildings predate NBE-CT-79 norm (1980)
    # Justification: No thermal insulation → 3-4x higher heat losses
    inefficient_building = pct_pre1980 > 60,
    
    # --------------------------------------------------------------------------
    # CRITERION 4: No urban mitigation
    # --------------------------------------------------------------------------
    # Threshold: NDVI < 0.12 (functional absence of vegetation)
    # Justification: No shade or evapotranspiration → +3-5°C temperature
    no_urban_mitigation = NDVI_mean_21 < 0.12,
    
    # --------------------------------------------------------------------------
    # FINAL TARGET: Physical energy vulnerability
    # --------------------------------------------------------------------------
    # Meets 3 or more of 4 criteria → high structural vulnerability
    n_physical_criteria_v2 = as.integer(extreme_winter) + 
                             as.integer(extreme_summer) + 
                             as.integer(inefficient_building) + 
                             as.integer(no_urban_mitigation),
    
    physical_vulnerability_v2 = n_physical_criteria_v2 >= 3
  )

# Descriptive statistics
cat("Physical vulnerability criteria distribution (V2 - no circularity):\n")
cat(sprintf("  - Extreme winter (LST_p10 < Q1):           %4d sections (%.1f%%)\n", 
            sum(dataset$extreme_winter, na.rm = TRUE),
            100 * mean(dataset$extreme_winter, na.rm = TRUE)))
cat(sprintf("  - Extreme summer (LST_p90 > 42°C):         %4d sections (%.1f%%)\n", 
            sum(dataset$extreme_summer, na.rm = TRUE),
            100 * mean(dataset$extreme_summer, na.rm = TRUE)))
cat(sprintf("  - Inefficient buildings (>60%% pre-1980):   %4d sections (%.1f%%)\n", 
            sum(dataset$inefficient_building, na.rm = TRUE),
            100 * mean(dataset$inefficient_building, na.rm = TRUE)))
cat(sprintf("  - No urban mitigation (NDVI < 0.12):       %4d sections (%.1f%%)\n", 
            sum(dataset$no_urban_mitigation, na.rm = TRUE),
            100 * mean(dataset$no_urban_mitigation, na.rm = TRUE)))

cat(sprintf("\n MODEL 2 TARGET (V2 - no circularity): %d vulnerable sections (%.1f%%)\n", 
            sum(dataset$physical_vulnerability_v2, na.rm = TRUE),
            100 * mean(dataset$physical_vulnerability_v2, na.rm = TRUE)))

# Distribution by number of criteria met
cat("\nDistribution by number of criteria met:\n")
criteria_table_v2 <- dataset |> 
  count(n_physical_criteria_v2) |> 
  mutate(percentage = 100 * n / sum(n))
print(criteria_table_v2, n = Inf)
```

### Step 5: Building the Model 1 Target — Energy Poverty

```{r}
# Target Model 1: EVI > 15 (energy poverty threshold)
dataset <- dataset |> 
  mutate(energy_poverty = EVI_synthetic > 15)

cat(sprintf(" MODEL 1 TARGET: %d sections with energy poverty (%.1f%%)\n", 
            sum(dataset$energy_poverty, na.rm = TRUE),
            100 * mean(dataset$energy_poverty, na.rm = TRUE)))

cat(sprintf("  Mean EVI: %.2f (SD: %.2f)\n", 
            mean(dataset$EVI_synthetic, na.rm = TRUE),
            sd(dataset$EVI_synthetic, na.rm = TRUE)))
cat(sprintf("  Median EVI: %.2f (Q1: %.2f, Q3: %.2f)\n",
            median(dataset$EVI_synthetic, na.rm = TRUE),
            quantile(dataset$EVI_synthetic, 0.25, na.rm = TRUE),
            quantile(dataset$EVI_synthetic, 0.75, na.rm = TRUE)))
```

### Step 6: Training Model 2 — Physical Vulnerability

#### Part 1: Predictor selection (excluding target components)

```{r}
# CRITICAL: Exclude variables directly used in target construction
predictors_m2_v2 <- c(
  # LST (EXCEPT winter_p10 and summer_p90, which are in the target)
  "LST_winter_mean_21", "LST_winter_sd_21", 
  "LST_winter_median_21", "LST_winter_p90_21",
  "LST_summer_mean_21", "LST_summer_sd_21", 
  "LST_summer_p10_21", "LST_summer_median_21",
  
  # NDVI (EXCEPT mean_21, used in target as threshold < 0.12)
  "NDVI_sd_21", "NDVI_p10_21", "NDVI_median_21", "NDVI_p90_21",
  
  # NDBI, ALBEDO, tcc
  "NDBI_mean_21", "NDBI_sd_21",
  "ALBEDO_mean_21", "ALBEDO_sd_21",
  "tcc",
  
  # Derived variables (allowed now - NOT in target)
  "seasonal_amplitude", "urban_heat_island"
)

predictors_m2_v2 <- predictors_m2_v2[predictors_m2_v2 %in% names(dataset)]

cat("Model 2 predictors (V2 - no circularity):\n")
cat(sprintf("  Total: %d variables\n", length(predictors_m2_v2)))
cat(sprintf("  - LST: %d variables\n", sum(grepl("LST_", predictors_m2_v2))))
cat(sprintf("  - NDVI: %d variables\n", sum(grepl("NDVI_", predictors_m2_v2))))
cat(sprintf("  - NDBI: %d variables\n", sum(grepl("NDBI_", predictors_m2_v2))))
cat(sprintf("  - ALBEDO: %d variables\n", sum(grepl("ALBEDO_", predictors_m2_v2))))
cat(sprintf("  - Derived: %d variables\n", sum(predictors_m2_v2 %in% c("seasonal_amplitude", "urban_heat_island"))))
```

#### Part 2: Train/Test split

```{r}
set.seed(42)
train_idx_m2_v2 <- createDataPartition(dataset$physical_vulnerability_v2, 
                                       p = 0.8, list = FALSE)

train_m2_v2 <- dataset[train_idx_m2_v2, ]
test_m2_v2  <- dataset[-train_idx_m2_v2, ]

cat(sprintf(" Train set: %d sections (%.1f%% vulnerable)\n", 
            nrow(train_m2_v2),
            100 * mean(train_m2_v2$physical_vulnerability_v2)))
cat(sprintf(" Test set:  %d sections (%.1f%% vulnerable)\n", 
            nrow(test_m2_v2),
            100 * mean(test_m2_v2$physical_vulnerability_v2)))
```


#### Part 3: Train random forest

```{r}
# CRITICAL: Convert target to factor to force classification
train_m2_v2 <- train_m2_v2 |> 
  mutate(physical_vulnerability_v2 = factor(physical_vulnerability_v2, levels = c(FALSE, TRUE)))

test_m2_v2 <- test_m2_v2 |> 
  mutate(physical_vulnerability_v2 = factor(physical_vulnerability_v2, levels = c(FALSE, TRUE)))

formula_m2_v2 <- as.formula(paste("physical_vulnerability_v2 ~", 
                                   paste(predictors_m2_v2, collapse = " + ")))

cat("Formula:", deparse(formula_m2_v2), "\n")
cat("Target levels:", levels(train_m2_v2$physical_vulnerability_v2), "\n")
cat("Train distribution:", table(train_m2_v2$physical_vulnerability_v2), "\n")

start_m2_v2 <- Sys.time()

rf_model_2_v2 <- randomForest(
  formula    = formula_m2_v2,
  data       = train_m2_v2,
  ntree      = 500,
  mtry       = floor(sqrt(length(predictors_m2_v2))),
  importance = TRUE,
  na.action  = na.omit
)

elapsed_m2_v2 <- as.numeric(difftime(Sys.time(), start_m2_v2, units = "secs"))

cat(sprintf(" Model 2 (V2) trained in %.1f seconds (CLASSIFICATION)\n", elapsed_m2_v2))
```

#### Part 4: evaluation

```{r}
# Predictions
pred_m2_v2_train <- predict(rf_model_2_v2, train_m2_v2, type = "response")
pred_m2_v2_test  <- predict(rf_model_2_v2, test_m2_v2,  type = "response")

# Probabilities (second column = TRUE class)
prob_matrix_train_v2 <- predict(rf_model_2_v2, train_m2_v2, type = "prob")
prob_matrix_test_v2  <- predict(rf_model_2_v2, test_m2_v2,  type = "prob")

pred_prob_m2_v2_train <- prob_matrix_train_v2[, 2]
pred_prob_m2_v2_test  <- prob_matrix_test_v2[, 2]

# Confusion matrices
cm_m2_v2_train <- confusionMatrix(
  factor(pred_m2_v2_train, levels = c("FALSE", "TRUE")),
  factor(train_m2_v2$physical_vulnerability_v2, levels = c("FALSE", "TRUE")),
  positive = "TRUE"
)

cm_m2_v2_test <- confusionMatrix(
  factor(pred_m2_v2_test, levels = c("FALSE", "TRUE")),
  factor(test_m2_v2$physical_vulnerability_v2, levels = c("FALSE", "TRUE")),
  positive = "TRUE"
)

cat("MODEL 2 RESULTS (V2 - NO CIRCULARITY):\n")
cat(strrep("-", 79), "\n")
cat(sprintf("TRAIN - Accuracy: %.3f | Precision: %.3f | Recall: %.3f | F1: %.3f\n",
            cm_m2_v2_train$overall['Accuracy'],
            cm_m2_v2_train$byClass['Precision'],
            cm_m2_v2_train$byClass['Recall'],
            cm_m2_v2_train$byClass['F1']))
cat(sprintf("TEST  - Accuracy: %.3f | Precision: %.3f | Recall: %.3f | F1: %.3f\n",
            cm_m2_v2_test$overall['Accuracy'],
            cm_m2_v2_test$byClass['Precision'],
            cm_m2_v2_test$byClass['Recall'],
            cm_m2_v2_test$byClass['F1']))

# AUC
roc_m2_v2 <- roc(as.numeric(test_m2_v2$physical_vulnerability_v2),
                 pred_prob_m2_v2_test, quiet = TRUE)
cat(sprintf("TEST  - AUC: %.3f\n", auc(roc_m2_v2)))

# Save metrics
metrics_m2_v2 <- data.frame(
  Set       = c("Train", "Test"),
  Accuracy  = c(cm_m2_v2_train$overall['Accuracy'], cm_m2_v2_test$overall['Accuracy']),
  Precision = c(cm_m2_v2_train$byClass['Precision'], cm_m2_v2_test$byClass['Precision']),
  Recall    = c(cm_m2_v2_train$byClass['Recall'],   cm_m2_v2_test$byClass['Recall']),
  F1        = c(cm_m2_v2_train$byClass['F1'],       cm_m2_v2_test$byClass['F1']),
  AUC       = c(NA, auc(roc_m2_v2))
)

metrics_m2_v2
```


#### Part 5: Variable importance

```{r}
importance_m2_v2 <- importance(rf_model_2_v2)
importance_m2_v2_df <- data.frame(
  Variable = rownames(importance_m2_v2),
  MeanDecreaseGini = importance_m2_v2[, "MeanDecreaseGini"]
) |> 
  arrange(desc(MeanDecreaseGini)) |> 
  mutate(Relative_Importance = 100 * MeanDecreaseGini / sum(MeanDecreaseGini))

cat("MOST IMPORTANT VARIABLES (Top 10):\n")
print(head(importance_m2_v2_df, 10), row.names = FALSE, digits = 3)

# Importance plot
importance_m2_v2_df |> 
  head(10) |> 
  ggplot(aes(x = reorder(Variable, Relative_Importance), 
             y = Relative_Importance)) +
  geom_col(fill = "#2E75B6", alpha = 0.8) +
  coord_flip() +
  labs(title = "Model 2 (V2): Most Important Variables",
       subtitle = "Physical vulnerability prediction (no circularity)",
       x = "", y = "Relative importance (%)") +
  theme_minimal(base_size = 14) +
  theme(plot.title = element_text(face = "bold"))
```

### Step 7: Training Model 1 — Energy Poverty

#### Part 1: Predictor selection

```{r}
predictors_m1 <- c(
  # GEE raw (ALL)
  "LST_winter_mean_21", "LST_winter_sd_21", "LST_winter_p10_21", 
  "LST_winter_median_21", "LST_winter_p90_21",
  "LST_summer_mean_21", "LST_summer_sd_21", "LST_summer_p10_21", 
  "LST_summer_median_21", "LST_summer_p90_21",
  "NDVI_mean_21", "NDVI_sd_21", "NDVI_p10_21", "NDVI_median_21", "NDVI_p90_21",
  "NDBI_mean_21", "NDBI_sd_21",
  "ALBEDO_mean_21", "ALBEDO_sd_21",
  "tcc",
  # Building age
  "mean_building_age", "pct_pre1980",
  # Derived climate variables
  "heating_demand", "cooling_demand", "seasonal_amplitude", "urban_heat_island"
)

predictors_m1 <- predictors_m1[predictors_m1 %in% names(dataset)]
```

#### Part 2: Train/Test split

```{r}
set.seed(42)
train_idx_m1 <- createDataPartition(dataset$energy_poverty, p = 0.8, list = FALSE)

train_m1 <- dataset[train_idx_m1, ]
test_m1  <- dataset[-train_idx_m1, ]

cat(sprintf(" Train set: %d sections (%.1f%% energy poor)\n", 
            nrow(train_m1),
            100 * mean(train_m1$energy_poverty)))
cat(sprintf(" Test set:  %d sections (%.1f%% energy poor)\n", 
            nrow(test_m1),
            100 * mean(test_m1$energy_poverty)))
```

#### Part 3: Train Random Forest

```{r}
train_m1 <- train_m1 |> mutate(energy_poverty = factor(energy_poverty, levels = c(FALSE, TRUE)))
test_m1  <- test_m1  |> mutate(energy_poverty = factor(energy_poverty, levels = c(FALSE, TRUE)))

formula_m1 <- as.formula(paste("energy_poverty ~", paste(predictors_m1, collapse = " + ")))

cat("Target levels:", levels(train_m1$energy_poverty), "\n")
cat("Train distribution:", table(train_m1$energy_poverty), "\n")

start_m1 <- Sys.time()

rf_model_1 <- randomForest(
  formula    = formula_m1,
  data       = train_m1,
  ntree      = 500,
  mtry       = floor(sqrt(length(predictors_m1))),
  importance = TRUE,
  na.action  = na.omit
)

elapsed_m1 <- as.numeric(difftime(Sys.time(), start_m1, units = "secs"))
```

#### Part 4: evaluation

```{r}
pred_m1_train <- predict(rf_model_1, train_m1, type = "response")
pred_m1_test  <- predict(rf_model_1, test_m1,  type = "response")

prob_matrix_train <- predict(rf_model_1, train_m1, type = "prob")
prob_matrix_test  <- predict(rf_model_1, test_m1,  type = "prob")

pred_prob_m1_train <- prob_matrix_train[, 2]
pred_prob_m1_test  <- prob_matrix_test[, 2]

cm_m1_train <- confusionMatrix(
  factor(pred_m1_train, levels = c("FALSE", "TRUE")),
  factor(train_m1$energy_poverty, levels = c("FALSE", "TRUE")),
  positive = "TRUE"
)

cm_m1_test <- confusionMatrix(
  factor(pred_m1_test, levels = c("FALSE", "TRUE")),
  factor(test_m1$energy_poverty, levels = c("FALSE", "TRUE")),
  positive = "TRUE"
)

cat("MODEL 1 RESULTS:\n")
cat(strrep("-", 79), "\n")
cat(sprintf("TRAIN - Accuracy: %.3f | Precision: %.3f | Recall: %.3f | F1: %.3f\n",
            cm_m1_train$overall['Accuracy'],
            cm_m1_train$byClass['Precision'],
            cm_m1_train$byClass['Recall'],
            cm_m1_train$byClass['F1']))
cat(sprintf("TEST  - Accuracy: %.3f | Precision: %.3f | Recall: %.3f | F1: %.3f\n",
            cm_m1_test$overall['Accuracy'],
            cm_m1_test$byClass['Precision'],
            cm_m1_test$byClass['Recall'],
            cm_m1_test$byClass['F1']))

roc_m1 <- roc(as.numeric(test_m1$energy_poverty), pred_prob_m1_test, quiet = TRUE)
cat(sprintf("TEST  - AUC: %.3f\n", auc(roc_m1)))

metrics_m1 <- data.frame(
  Set       = c("Train", "Test"),
  Accuracy  = c(cm_m1_train$overall['Accuracy'], cm_m1_test$overall['Accuracy']),
  Precision = c(cm_m1_train$byClass['Precision'], cm_m1_test$byClass['Precision']),
  Recall    = c(cm_m1_train$byClass['Recall'],   cm_m1_test$byClass['Recall']),
  F1        = c(cm_m1_train$byClass['F1'],       cm_m1_test$byClass['F1']),
  AUC       = c(NA, auc(roc_m1))
)

metrics_m1
```

#### Variable importance

```{r}
importance_m1 <- importance(rf_model_1)
importance_m1_df <- data.frame(
  Variable = rownames(importance_m1),
  MeanDecreaseGini = importance_m1[, "MeanDecreaseGini"]
) |> 
  arrange(desc(MeanDecreaseGini)) |> 
  mutate(Relative_Importance = 100 * MeanDecreaseGini / sum(MeanDecreaseGini))

cat("MOST IMPORTANT VARIABLES (Top 10):\n")
print(head(importance_m1_df, 10), row.names = FALSE, digits = 3)

importance_m1_df |> 
  head(10) |> 
  ggplot(aes(x = reorder(Variable, Relative_Importance), 
             y = Relative_Importance)) +
  geom_col(fill = "#C00000", alpha = 0.8) +
  coord_flip() +
  labs(title = "Model 1: Most Important Variables",
       subtitle = "Energy poverty prediction (EVI > 15)",
       x = "", y = "Relative importance (%)") +
  theme_minimal(base_size = 14) +
  theme(plot.title = element_text(face = "bold"))
dev.off()
```

### Step 8: Group identification (model 1 vs sae)

```{r}
# Predict on the full dataset
dataset <- dataset |> 
  mutate(
    energy_poverty = factor(energy_poverty, levels = c(FALSE, TRUE)),
    physical_vulnerability_v2 = factor(physical_vulnerability_v2, levels = c(FALSE, TRUE))
  ) |> 
  mutate(
    pred_prob_m1    = predict(rf_model_1,    dataset, type = "prob")[, 2],
    pred_m1         = predict(rf_model_1,    dataset, type = "response") == "TRUE",
    pred_prob_m2_v2 = predict(rf_model_2_v2, dataset, type = "prob")[, 2],
    pred_m2_v2      = predict(rf_model_2_v2, dataset, type = "response") == "TRUE"
  )

cat("8.1. Classifying sections into groups...\n\n")

dataset <- dataset |> 
  mutate(
    group = case_when(
      # CONCORDANT: both indicate vulnerable
      pred_m1 == TRUE  & EVI_synthetic >= 15 ~ "Concordant",
      
      # SPATIAL FALSE POSITIVES: model says YES, SAE says NO
      pred_m1 == TRUE  & EVI_synthetic <  15 ~ "Spatial_False_Positives",
      
      # SPATIAL FALSE NEGATIVES: model says NO, SAE says YES
      pred_m1 == FALSE & EVI_synthetic >= 15 ~ "Spatial_False_Negatives",
      
      # NON-VULNERABLE: both indicate non-vulnerable
      pred_m1 == FALSE & EVI_synthetic <  15 ~ "Non_Vulnerable",
      
      TRUE ~ "Other"
    )
  )

# Group distribution
group_table <- dataset |> 
  count(group) |> 
  mutate(percentage = 100 * n / sum(n)) |> 
  arrange(desc(n))

cat("GROUP DISTRIBUTION:\n")
print(group_table, row.names = FALSE)

cat(sprintf("\n Spatial False Positives: %d sections (%.1f%% of total)\n", 
            sum(dataset$group == "Spatial_False_Positives"),
            100 * mean(dataset$group == "Spatial_False_Positives")))
cat(sprintf("Spatial False Negatives: %d sections (%.1f%% of total)\n", 
            sum(dataset$group == "Spatial_False_Negatives"),
            100 * mean(dataset$group == "Spatial_False_Negatives")))
```

### Step 9: Group characterization

```{r}
# Compare: Spatial False Positives vs Concordant
comparison <- dataset |> 
  filter(group %in% c("Spatial_False_Positives", "Concordant")) |> 
  group_by(group) |> 
  summarise(
    n = n(),
    # Economic
    mean_EVI = mean(EVI_synthetic, na.rm = TRUE),
    sd_EVI   = sd(EVI_synthetic, na.rm = TRUE),
    mean_income = mean(mean_equiv_income, na.rm = TRUE),
    sd_income   = sd(mean_equiv_income, na.rm = TRUE),
    # Physical vulnerability (Model 2 V2)
    pct_phys_vuln_m2_v2 = 100 * mean(as.logical(physical_vulnerability_v2), na.rm = TRUE),
    mean_n_physical_criteria = mean(n_physical_criteria_v2, na.rm = TRUE),
    # Specific criteria
    pct_extreme_winter      = 100 * mean(extreme_winter, na.rm = TRUE),
    pct_extreme_summer      = 100 * mean(extreme_summer, na.rm = TRUE),
    pct_inefficient_bldg    = 100 * mean(inefficient_building, na.rm = TRUE),
    pct_no_urban_mitigation = 100 * mean(no_urban_mitigation, na.rm = TRUE),
    # Physical variables
    mean_heating_demand = mean(heating_demand, na.rm = TRUE),
    mean_cooling_demand = mean(cooling_demand, na.rm = TRUE),
    mean_LST_winter_p10 = mean(LST_winter_p10_21, na.rm = TRUE),
    mean_LST_summer_p90 = mean(LST_summer_p90_21, na.rm = TRUE),
    mean_NDVI           = mean(NDVI_mean_21, na.rm = TRUE),
    mean_building_age   = mean(mean_building_age, na.rm = TRUE),
    mean_pct_pre1980    = mean(pct_pre1980, na.rm = TRUE),
    .groups = "drop"
  ) |> 
  pivot_longer(-c(group, n), names_to = "variable", values_to = "value") |> 
  pivot_wider(names_from = group, values_from = value) |> 
  mutate(
    Difference          = Spatial_False_Positives - Concordant,
    Relative_Diff_Pct   = 100 * (Spatial_False_Positives - Concordant) / Concordant
  )

# Statistical tests
cat("\n9.1. Statistical tests of differences...\n\n")

false_pos   <- dataset |> filter(group == "Spatial_False_Positives")
concordant  <- dataset |> filter(group == "Concordant")

cat("STATISTICAL TESTS (Spatial False Positives vs Concordant):\n")
cat(strrep("=", 79), "\n")

test_vars <- c("EVI_synthetic", "mean_equiv_income", "heating_demand", 
               "cooling_demand", "NDVI_mean_21", "mean_building_age")

for (var in test_vars) {
  if (var %in% names(dataset)) {
    test <- t.test(false_pos[[var]], concordant[[var]])
    cat(sprintf("\n%s:\n", var))
    cat(sprintf("  Spatial False Positives: %.2f (SD: %.2f)\n", 
                mean(false_pos[[var]], na.rm = TRUE), 
                sd(false_pos[[var]], na.rm = TRUE)))
    cat(sprintf("  Concordant:              %.2f (SD: %.2f)\n", 
                mean(concordant[[var]], na.rm = TRUE), 
                sd(concordant[[var]], na.rm = TRUE)))
    cat(sprintf("  Difference:              %.2f\n", 
                mean(false_pos[[var]], na.rm = TRUE) - mean(concordant[[var]], na.rm = TRUE)))
    cat(sprintf("  p-value:                 %.4f %s\n", test$p.value, 
                ifelse(test$p.value < 0.05, "***", "")))
  }
}
```

### Step 10: visualizations

```{r}
# ROC curves
plot(roc_m1, main = sprintf("Model 1: Energy Poverty\nAUC = %.3f", auc(roc_m1)),
     col = "#C00000", lwd = 2)

plot(roc_m2_v2, main = sprintf("Model 2 (V2): Physical Vulnerability\nAUC = %.3f", auc(roc_m2_v2)),
     col = "#2E75B6", lwd = 2)



# EVI distribution by group

dataset |> 
  filter(group %in% c("Spatial_False_Positives", "Concordant", "Non_Vulnerable")) |> 
  ggplot(aes(x = group, y = EVI_synthetic, fill = group)) +
  geom_boxplot(alpha = 0.7) +
  geom_hline(yintercept = 15, linetype = "dashed", color = "red", size = 1) +
  scale_fill_manual(values = c("Spatial_False_Positives" = "#FFA500", 
                                "Concordant" = "#C00000",
                                "Non_Vulnerable" = "#2E75B6")) +
  labs(title = "EVI Distribution by Group",
       subtitle = "Red line = energy poverty threshold (EVI = 15)",
       x = "", y = "EVI synthetic") +
  theme_minimal(base_size = 14) +
  theme(legend.position = "none",
        plot.title = element_text(face = "bold"),
        axis.text.x = element_text(angle = 20, hjust = 1))


# Scatter: Income vs EVI

dataset |> 
  filter(group %in% c("Spatial_False_Positives", "Concordant", "Non_Vulnerable")) |> 
  ggplot(aes(x = mean_equiv_income, y = EVI_synthetic, color = group)) +
  geom_point(alpha = 0.6, size = 2) +
  geom_hline(yintercept = 15, linetype = "dashed", color = "red") +
  geom_vline(xintercept = 16000, linetype = "dashed", color = "blue") +
  scale_color_manual(values = c("Spatial_False_Positives" = "#FFA500", 
                                 "Concordant" = "#C00000",
                                 "Non_Vulnerable" = "#2E75B6")) +
  labs(title = "Income vs EVI by Group",
       subtitle = "Red line = EVI 15 | Blue line = Income ~16k€ (poverty threshold)",
       x = "Mean income (€)", y = "EVI synthetic", color = "Group") +
  theme_minimal(base_size = 14) +
  theme(plot.title = element_text(face = "bold"))

dev.off()
```

### Conclusion

Phase 3 demonstrates that **Google Earth Engine variables, when properly 
combined with building-age data, can recover much of the socioeconomic 
classification produced by the SAE model**. Both classifiers achieve high 
discriminative ability (AUC > 0.9), but they capture **partially different 
phenomena**:

- Model 1 (energy poverty) recovers the *socioeconomic* dimension of 
  vulnerability inferred from income, household composition, and building 
  characteristics.
- Model 2 (physical vulnerability) captures *structural* adverse conditions 
  visible from satellite imagery, independent of income.

The **discrepancy analysis** between Model 1 and the SAE classification 
identifies sections that are *invisible* to traditional survey-based targeting:

- **Spatial False Positives** (≈ 1.4%) — wealthy areas with adverse physical 
  conditions, where economic capacity neutralizes structural risks.
- **Spatial False Negatives** (≈ 1.7%) — low-income areas with relatively 
  good physical infrastructure, where economic vulnerability dominates over 
  physical conditions.

These findings support a **dual-targeting strategy** for energy poverty 
intervention, combining survey-based socioeconomic indicators with 
satellite-derived environmental signals. Neither source alone is sufficient — 
their intersection identifies the most acute cases (Concordant), while their 
discrepancies reveal policy-relevant population subgroups that single-source 
methods systematically miss.

The methodological correction from V1 to V2 (removing circularity) is 
essential: the apparent perfect fit of V1 was an artifact of using target-defining 
variables as predictors. The V2 metrics (Accuracy ≈ 93%, AUC ≈ 0.965) are 
realistic and reflect genuine predictive capacity, supporting the use of GEE 
data for this type of urban vulnerability mapping.



## Shap Values

```{r}
# =============================================================================
# SHAP VALUES — Local interpretability (using iml package)
# =============================================================================

library(iml)

cat("\n")
cat(strrep("=", 79), "\n")
cat("SHAP VALUES — LOCAL FEATURE CONTRIBUTIONS\n")
cat(strrep("=", 79), "\n\n")

# Prepare data
set.seed(42)
sample_idx <- sample(1:nrow(test_m2_v2), min(200, nrow(test_m2_v2)))
shap_sample <- test_m2_v2[sample_idx, predictors_m2_v2]

# Create predictor object (wrapper around your trained model)
predictor_m2 <- Predictor$new(
  model = rf_model_2_v2,
  data = shap_sample,
  y = test_m2_v2$physical_vulnerability_v2[sample_idx],
  type = "prob",
  class = "TRUE"
)

cat("Computing SHAP values for Model 2 (200 sections)...\n")

# Compute SHAP for ALL sample observations
shap_results <- list()
for (i in 1:nrow(shap_sample)) {
  if (i %% 50 == 0) cat(sprintf("  Processed %d/%d sections\n", i, nrow(shap_sample)))
  
  shapley <- Shapley$new(predictor_m2, x.interest = shap_sample[i, ])
  shap_results[[i]] <- shapley$results
}

cat("✓ SHAP values computed\n\n")

# Combine results into a data frame
shap_df <- do.call(rbind, lapply(seq_along(shap_results), function(i) {
  shap_results[[i]] |> 
    mutate(observation = i)
}))

# Global SHAP summary
shap_summary_m2 <- shap_df |> 
  group_by(feature) |> 
  summarise(
    Mean_Abs_SHAP = mean(abs(phi)),
    Mean_SHAP = mean(phi),
    .groups = "drop"
  ) |> 
  arrange(desc(Mean_Abs_SHAP)) |> 
  rename(Variable = feature)

cat("TOP 10 VARIABLES BY SHAP IMPORTANCE (Model 2):\n")
print(head(shap_summary_m2, 10), row.names = FALSE, digits = 3)

# Visualization
shap_summary_m2 |> 
  head(15) |> 
  ggplot(aes(x = reorder(Variable, Mean_Abs_SHAP), y = Mean_Abs_SHAP)) +
  geom_col(aes(fill = Mean_SHAP > 0), alpha = 0.8) +
  coord_flip() +
  scale_fill_manual(values = c("TRUE" = "#C00000", "FALSE" = "#2E75B6"),
                    labels = c("TRUE" = "Pushes toward vulnerable", 
                               "FALSE" = "Pushes toward non-vulnerable"),
                    name = "") +
  labs(title = "SHAP Feature Importance — Model 2",
       subtitle = "Mean absolute SHAP value (impact on prediction)",
       x = "", y = "Mean |SHAP value|") +
  theme_minimal(base_size = 13) +
  theme(plot.title = element_text(face = "bold"),
        legend.position = "top")
```

### Final summary

```{r}
cat("MODEL 1 — Energy Poverty Prediction (EVI > 15):\n")
cat(sprintf("  Accuracy: %.1f%%\n", 100 * cm_m1_test$overall['Accuracy']))
cat(sprintf("  AUC: %.3f\n", auc(roc_m1)))
cat(sprintf("  F1-Score: %.3f\n", cm_m1_test$byClass['F1']))
cat(sprintf("  Top variable: %s (%.1f%%)\n", 
            importance_m1_df$Variable[1], 
            importance_m1_df$Relative_Importance[1]))

cat("\nMODEL 2 (V2) — Physical Vulnerability Prediction (no circularity):\n")
cat(sprintf("  Accuracy: %.1f%%\n", 100 * cm_m2_v2_test$overall['Accuracy']))
cat(sprintf("  AUC: %.3f\n", auc(roc_m2_v2)))
cat(sprintf("  F1-Score: %.3f\n", cm_m2_v2_test$byClass['F1']))
cat(sprintf("  Top variable (MeanDecreaseGini): %s (%.1f%%)\n", 
            importance_m2_v2_df$Variable[1], 
            importance_m2_v2_df$Relative_Importance[1]))
cat(sprintf("  Top variable (SHAP):             %s (%.3f)\n", 
            shap_summary_m2$Variable[1], 
            shap_summary_m2$Mean_Abs_SHAP[1]))

cat("\nGROUP DISTRIBUTION:\n")
print(group_table, row.names = FALSE)
```

## Results

This section presents the empirical findings of the four-phase methodology 
applied to the 2,443 census sections of Madrid. Results are organised in four 
subsections: (1) spatial distribution of energy vulnerability; (2) predictive 
performance of the dual Random Forest; (3) discrepancy analysis; and (4) 
statistical characterization of the discrepancy groups.

### 1. Spatial Distribution of Energy Vulnerability

```{r}
# Table 1 — External validation: correlations with INE Atlas indicators
corr_table <- data.frame(
  Indicator = c("Mean equivalised income (€)",
                "% population below 60% threshold",
                "Gini coefficient",
                "% dwellings pre-1980"),
  Pearson_r = c(correlations$cor_income,
                correlations$cor_threshold60,
                correlations$cor_gini,
                correlations$cor_pre1980),
  Expected_sign = c("negative", "positive", "positive", "positive"),
  Strength = c(
    ifelse(abs(correlations$cor_income) > 0.7, "Strong",
           ifelse(abs(correlations$cor_income) > 0.4, "Moderate", "Weak")),
    ifelse(abs(correlations$cor_threshold60) > 0.7, "Strong",
           ifelse(abs(correlations$cor_threshold60) > 0.4, "Moderate", "Weak")),
    ifelse(abs(correlations$cor_gini) > 0.7, "Strong",
           ifelse(abs(correlations$cor_gini) > 0.4, "Moderate", "Weak")),
    ifelse(abs(correlations$cor_pre1980) > 0.7, "Strong",
           ifelse(abs(correlations$cor_pre1980) > 0.4, "Moderate", "Weak"))
  )
)

flextable(corr_table) |>
  set_header_labels(
    Indicator     = "INE Atlas indicator",
    Pearson_r     = "Pearson r",
    Expected_sign = "Expected sign",
    Strength      = "Strength"
  ) |>
  colformat_double(j = "Pearson_r", digits = 3) |>
  bold(part = "header") |>
  bg(part = "header", bg = "#2E75B6") |>
  color(part = "header", color = "white") |>
  align(align = "left",   j = "Indicator") |>
  align(align = "center", j = c("Pearson_r", "Expected_sign", "Strength")) |>
  autofit() |>
  set_caption("Table 1 — External validation of the synthetic EVI: Pearson correlations between section-level EVI estimates and four independent INE Atlas indicators not used in the SAE model (n = 2,443 census sections). Source: author's elaboration.")
```

```{r}
# Figure 1 — External validation: bivariate relationships
p_income_v2 <- ggplot(sae_results, aes(x = income * 10000, y = EVI_synthetic)) +
  geom_point(alpha = 0.4, color = "#457B9D", size = 1) +
  geom_smooth(method = "loess", color = "#E63946", se = TRUE, linewidth = 1.2) +
  scale_x_continuous(labels = dollar_format(prefix = "€", big.mark = ",")) +
  labs(
    title = "A) EVI vs Mean Income per Equivalence Unit",
    subtitle = sprintf("Correlation: r = %.3f", correlations$cor_income),
    x = "Mean Income (€/EU)",
    y = "EVI (%)"
  ) +
  theme_classic(base_size = 10) +
  theme(plot.title    = element_text(face = "bold", size = 11),
        plot.subtitle = element_text(size = 9, color = "gray40"))

p_threshold_v2 <- ggplot(sae_results, aes(x = pct_below_threshold_60, y = EVI_synthetic)) +
  geom_point(alpha = 0.4, color = "#457B9D", size = 1) +
  geom_smooth(method = "lm", color = "#E63946", se = TRUE, linewidth = 1.2) +
  labs(
    title = "B) EVI vs Population Below Poverty Threshold (60%)",
    subtitle = sprintf("Correlation: r = %.3f", correlations$cor_threshold60),
    x = "% population below 60% threshold",
    y = "EVI (%)"
  ) +
  theme_classic(base_size = 10) +
  theme(plot.title    = element_text(face = "bold", size = 11),
        plot.subtitle = element_text(size = 9, color = "gray40"))

fig_1 <- p_income_v2 | p_threshold_v2
fig_1
```
*Figure 1: External validation of the synthetic EVI: non-linear inverse relationship 
with mean equivalised income (Panel A, r = −0.900) and linear positive relationship 
with the percentage of population below the 60% poverty threshold (Panel B, r = 0.924). 
Source: author's elaboration based on the Synthetic Regression Estimator (Phase 2) and INE Atlas 2021.*


```{r}
# Figures 3 and 4 — Choropleth maps
madrid_sf <- st_read("data/seccionado_2021/SECC_CE_20210101.shp") |>
  filter(str_detect(CUSEC, "^28079"))

madrid_map <- madrid_sf |>
  left_join(
    dataset |> select(CUSEC, EVI_synthetic, energy_poverty, group, mean_equiv_income),
    by = "CUSEC"
  ) |>
  mutate(
    district = str_sub(CUSEC, 6, 7),
    group_label = case_when(
      group == "Concordant"              ~ "Concordant",
      group == "Non_Vulnerable"          ~ "Non-vulnerable",
      group == "Spatial_False_Positives" ~ "Spatial False Positives",
      group == "Spatial_False_Negatives" ~ "Spatial False Negatives"
    ),
    group_label = factor(group_label, levels = c(
      "Concordant", "Spatial False Positives",
      "Spatial False Negatives", "Non-vulnerable"
    ))
  )

district_sf <- madrid_map |>
  group_by(district) |>
  summarise(geometry = st_union(geometry), .groups = "drop")

# Figure 3 — Continuous EVI map
map1 <- ggplot(madrid_map) +
  geom_sf(aes(fill = EVI_synthetic), color = NA) +
  geom_sf(data = district_sf, fill = NA, color = "white", linewidth = 0.4) +
  scale_fill_gradientn(
    colours = c("#2166AC", "#92C5DE", "#F7F7F7", "#F4A582", "#B2182B"),
    name = "Synthetic EVI (%)", na.value = "grey80"
  ) +
  labs(title = "Synthetic Energy Vulnerability Index",
       subtitle = "Continuous EVI by census section (n = 2,443) | Madrid 2021") +
  theme_void(base_size = 11) +
  theme(plot.title    = element_text(face = "bold", size = 13),
        plot.subtitle = element_text(color = "gray40", size = 10),
        legend.position = "right")
map1
```
*Figure 3: Spatial distribution of the Synthetic Energy Vulnerability Index (EVI) across 
Madrid's 2,443 census sections. Colour gradient from blue (low vulnerability) to red 
(high vulnerability). White boundaries delimit the 21 municipal districts. Source: own 
elaboration based on the Synthetic Regression Estimator (Phase 2), INE Atlas of Income 
Distribution 2021, and INE Census Section Cartography 2021 (SECC_CE_20210101).*

```{r}
# Figure 4 — Binary classification map
map2 <- ggplot(madrid_map) +
  geom_sf(aes(fill = energy_poverty), color = NA) +
  geom_sf(data = district_sf, fill = NA, color = "white", linewidth = 0.4) +
  scale_fill_manual(
    values = c("TRUE" = "#B2182B", "FALSE" = "#2166AC"),
    labels = c("TRUE" = "Energy poor (EVI > 15)", "FALSE" = "Non-poor"),
    name = NULL, na.value = "grey80"
  ) +
  labs(title = "Binary Energy Poverty Classification",
       subtitle = "EVI > 15 threshold | 40.0% of sections classified as energy poor") +
  theme_void(base_size = 11) +
  theme(plot.title    = element_text(face = "bold", size = 13),
        plot.subtitle = element_text(color = "gray40", size = 10),
        legend.position = "bottom")
map2
```
*Figure 4: Binary classification of Madrid's 2,443 census sections by energy poverty 
status (EVI > 15 threshold). Red sections (40.0%, n = 977) are classified as energy poor; 
blue sections (60.0%, n = 1,466) fall below the vulnerability threshold. White boundaries 
delimit the 21 municipal districts. Source: own elaboration based on the Synthetic 
Regression Estimator (Phase 2), INE Atlas of Income Distribution 2021, and INE Census 
Section Cartography 2021 (SECC_CE_20210101).*

### 2. Predictive Performance of the Dual Random Forest

```{r}
# Table 2 — Performance metrics
metrics_comparison <- data.frame(
  Metric = c("Accuracy", "Precision", "Recall", "F1-score", "AUC"),
  `Model 1 (Energy Poverty)` = c(
    sprintf("%.3f", cm_m1_test$overall['Accuracy']),
    sprintf("%.3f", cm_m1_test$byClass['Precision']),
    sprintf("%.3f", cm_m1_test$byClass['Recall']),
    sprintf("%.3f", cm_m1_test$byClass['F1']),
    sprintf("%.3f", auc(roc_m1))
  ),
  `Model 2 (Physical Vulnerability)` = c(
    sprintf("%.3f", cm_m2_v2_test$overall['Accuracy']),
    sprintf("%.3f", cm_m2_v2_test$byClass['Precision']),
    sprintf("%.3f", cm_m2_v2_test$byClass['Recall']),
    sprintf("%.3f", cm_m2_v2_test$byClass['F1']),
    sprintf("%.3f", auc(roc_m2_v2))
  ),
  check.names = FALSE
)

metrics_comparison |>
  flextable() |>
  bold(part = "header") |>
  bg(part = "header", bg = "#2E75B6") |>
  color(part = "header", color = "white") |>
  align(align = "left",   j = "Metric") |>
  align(align = "center", j = c(2, 3)) |>
  bold(j = "Metric") |>
  autofit() |>
  set_caption("Table 2 — Test-set performance metrics of the dual Random Forest classifiers. Model 1 predicts energy poverty (EVI > 15) using environmental and building age variables; Model 2 predicts physical vulnerability using exclusively environmental predictors. Source: author's elaboration based on Random Forest models trained on 80% of the sample and evaluated on the held-out 20% (n ≈ 489 sections).")
```

```{r}
# Figure 5 — Variable importance rankings
clean_var_names <- function(x) {
  x |> str_replace_all("_", " ") |> str_replace("21", "") |> str_trim()
}

top_m1 <- importance_m1_df |>
  head(10) |>
  mutate(Variable_clean = clean_var_names(Variable))

p_m1_imp <- ggplot(top_m1,
                   aes(x = reorder(Variable_clean, Relative_Importance),
                       y = Relative_Importance)) +
  geom_col(fill = "#C00000", alpha = 0.85) +
  geom_text(aes(label = sprintf("%.1f%%", Relative_Importance)),
            hjust = -0.2, size = 3.3) +
  coord_flip() +
  scale_y_continuous(limits = c(0, max(top_m1$Relative_Importance) * 1.15)) +
  labs(title = "A) Model 1 — Energy Poverty (EVI > 15)",
       subtitle = "Top 10 most important variables",
       x = NULL, y = "Relative importance (%)") +
  theme_classic(base_size = 10) +
  theme(plot.title    = element_text(face = "bold", size = 11),
        plot.subtitle = element_text(size = 9, color = "gray40"),
        axis.text.y   = element_text(size = 9))

top_m2 <- importance_m2_v2_df |>
  head(10) |>
  mutate(Variable_clean = clean_var_names(Variable))

p_m2_imp <- ggplot(top_m2,
                   aes(x = reorder(Variable_clean, Relative_Importance),
                       y = Relative_Importance)) +
  geom_col(fill = "#2E75B6", alpha = 0.85) +
  geom_text(aes(label = sprintf("%.1f%%", Relative_Importance)),
            hjust = -0.2, size = 3.3) +
  coord_flip() +
  scale_y_continuous(limits = c(0, max(top_m2$Relative_Importance) * 1.15)) +
  labs(title = "B) Model 2 — Physical Vulnerability",
       subtitle = "Top 10 most important variables (environmental only)",
       x = NULL, y = "Relative importance (%)") +
  theme_classic(base_size = 10) +
  theme(plot.title    = element_text(face = "bold", size = 11),
        plot.subtitle = element_text(size = 9, color = "gray40"),
        axis.text.y   = element_text(size = 9))

fig_5 <- p_m1_imp | p_m2_imp
fig_5
```
*Figure 5: Variable importance rankings of the dual Random Forest classifiers, measured 
by MeanDecreaseGini (Breiman, 2001). Panel A: top 10 most important predictors of Model 1 
(energy poverty). Panel B: top 10 most important predictors of Model 2 (physical 
vulnerability). Source: author's elaboration based on Random Forest models implemented 
with the randomForest R package (Liaw & Wiener, 2002) and Google Earth Engine satellite data.*

### 3. Discrepancy Analysis

```{r}
# Table 3 — Distribution of the four groups
group_table <- dataset |>
  count(group) |>
  mutate(percentage = 100 * n / sum(n)) |>
  arrange(desc(n))

table_3_data <- group_table |>
  mutate(
    Group_label = case_when(
      group == "Concordant"              ~ "Concordant (both vulnerable)",
      group == "Non_Vulnerable"          ~ "Non-vulnerable (both safe)",
      group == "Spatial_False_Positives" ~ "Spatial False Positives (RF vulnerable, SAE not)",
      group == "Spatial_False_Negatives" ~ "Spatial False Negatives (SAE vulnerable, RF not)"
    ),
    Interpretation = case_when(
      group == "Concordant"              ~ "Joint economic + physical vulnerability",
      group == "Non_Vulnerable"          ~ "No vulnerability detected",
      group == "Spatial_False_Positives" ~ "Adverse environment, adequate income",
      group == "Spatial_False_Negatives" ~ "Low income, adequate environment"
    )
  ) |>
  select(Group_label, n, percentage, Interpretation)

table_3_data |>
  flextable() |>
  set_header_labels(
    Group_label    = "Group",
    n              = "N sections",
    percentage     = "%",
    Interpretation = "Interpretation"
  ) |>
  colformat_double(j = "percentage", digits = 1, suffix = "%") |>
  colformat_int(j = "n", big.mark = ",") |>
  bold(part = "header") |>
  bg(part = "header", bg = "#2E75B6") |>
  color(part = "header", color = "white") |>
  align(align = "left",   j = c("Group_label", "Interpretation")) |>
  align(align = "center", j = c("n", "percentage")) |>
  bold(j = "Group_label") |>
  autofit() |>
  set_caption("Table 3 — Distribution of Madrid's 2,443 census sections across the four mutually exclusive groups defined by the intersection of the SAE-based classification (EVI > 15) and Model 1's environmental prediction. Source: author's elaboration based on the cross-classification of Phase 2 (SAE) and Phase 3 (Random Forest Model 1) outputs.")
```

```{r}
# Figure 6 — Discrepancy groups map
map3 <- ggplot(madrid_map) +
  geom_sf(aes(fill = group_label), color = NA) +
  geom_sf(data = district_sf, fill = NA, color = "white", linewidth = 0.4) +
  scale_fill_manual(
    values = c(
      "Concordant"               = "#E57373",
      "Spatial False Positives"  = "#FF8C00",
      "Spatial False Negatives"  = "#8E44AD",
      "Non-vulnerable"           = "#82B0D9"
    ),
    name = NULL, na.value = "grey80"
  ) +
  theme_void(base_size = 11) +
  theme(plot.title    = element_text(face = "bold", size = 13),
        plot.subtitle = element_text(color = "gray40", size = 10),
        legend.position = "bottom",
        legend.text = element_text(size = 9))
map3
```
*Figure 6: Spatial distribution of the four discrepancy groups across Madrid's 2,443 
census sections. Concordant (red, 38.2%) and Non-vulnerable (blue, 58.7%) sections 
dominate the territorial picture. Spatial False Positives (orange, 1.4%, n = 34) cluster 
in consolidated inner-city districts undergoing gentrification; Spatial False Negatives 
(purple, 1.7%, n = 42) are dispersed across peripheral areas. White boundaries delimit 
the 21 municipal districts. Source: own elaboration based on Phase 2 (SAE) and Phase 3 
(Random Forest Model 1) outputs, and INE Census Section Cartography 2021.*

### 4. Group Characterization and Statistical Validation

```{r}
# Table 4 — Statistical comparison: False Positives vs Concordant
false_pos  <- dataset |> filter(group == "Spatial_False_Positives")
concordant <- dataset |> filter(group == "Concordant")

vars_to_test <- c(
  "EVI_synthetic", "mean_equiv_income",
  "pct_pre1980", "mean_building_age",
  "heating_demand", "cooling_demand",
  "NDVI_mean_21", "urban_heat_island",
  "LST_summer_p90_21", "LST_winter_p10_21"
)

labels_vars <- c(
  "Synthetic EVI (%)", "Mean equivalised income (€)",
  "% dwellings pre-1980", "Mean building age (years)",
  "Heating demand", "Cooling demand",
  "NDVI mean (vegetation)", "Urban heat island index",
  "LST summer p90 (°C)", "LST winter p10 (°C)"
)

categories <- c(
  "Economic", "Economic",
  "Physical", "Physical", "Physical", "Physical",
  "Physical", "Physical", "Physical", "Physical"
)

results <- map_dfr(seq_along(vars_to_test), function(i) {
  var  <- vars_to_test[i]
  test <- t.test(false_pos[[var]], concordant[[var]])
  sig  <- ifelse(test$p.value < 0.001, "***",
          ifelse(test$p.value < 0.01,  "**",
          ifelse(test$p.value < 0.05,  "*", "ns")))
  data.frame(
    Category  = categories[i],
    Variable  = labels_vars[i],
    FP_mean   = mean(false_pos[[var]], na.rm = TRUE),
    FP_sd     = sd(false_pos[[var]],   na.rm = TRUE),
    Conc_mean = mean(concordant[[var]], na.rm = TRUE),
    Conc_sd   = sd(concordant[[var]],  na.rm = TRUE),
    p_value   = test$p.value,
    Sig       = sig
  )
})

results |>
  mutate(
    `False Positives (mean ± SD)` = sprintf("%.2f (±%.2f)", FP_mean, FP_sd),
    `Concordant (mean ± SD)`      = sprintf("%.2f (±%.2f)", Conc_mean, Conc_sd),
    `p-value` = case_when(
      p_value < 0.001 ~ "< 0.001 ***",
      p_value < 0.01  ~ sprintf("%.3f **", p_value),
      p_value < 0.05  ~ sprintf("%.3f *",  p_value),
      TRUE            ~ sprintf("%.3f",    p_value)
    )
  ) |>
  select(Category, Variable,
         `False Positives (mean ± SD)`,
         `Concordant (mean ± SD)`,
         `p-value`) |>
  flextable() |>
  bold(part = "header") |>
  bg(part = "header", bg = "#2E75B6") |>
  color(part = "header", color = "white") |>
  align(align = "left",   j = c("Category", "Variable")) |>
  align(align = "center", j = 3:5) |>
  bold(j = "Variable") |>
  bg(i = ~Category == "Economic", bg = "#FFF2CC") |>
  bg(i = ~Category == "Physical", bg = "#F2F2F2") |>
  merge_v(j = "Category") |>
  autofit() |>
  set_caption("Table 4 — Statistical comparison between Spatial False Positives (n = 34) and Concordant sections (n = 934) on ten key indicators of economic and physical vulnerability. Significance codes: *** p < 0.001, ** p < 0.01, * p < 0.05, ns: not significant. Source: author's elaboration based on Phase 3 outputs and the integrated dataset combining INE Atlas, Censo de Edificios CAM, and Google Earth Engine variables.")
```
